### 2024-06-24 report.Rmd
Write report

---
title: "DPM Simulation"
author: "Scott Claessens"
date: "`r Sys.Date()`"
bibliography: references.bib
output: 
  bookdown::pdf_document2:
    toc: false
    number_sections: false
---

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(knitr)
library(papaja)
library(posterior)
library(targets)
library(tidyverse)
```

# 2.3. Synthetic example

In order to demonstrate that our formal model is able to accurately recover true 
coevolutionary dynamics, we use a simple synthetic example.

Consider an ancestral species with two continuous traits, $X$ and $Y$. We 
simulate the coevolution of these traits over time, allowing traits to have 
autoregressive effects (_i.e._, change in a trait influences itself in the 
future), coevolutionary effects (_i.e._, change in a trait influences the other 
trait in the future), and stochastic drift. In this particular simulation, we 
specify that changes in $X$ influence future changes in $Y$, but not vice versa.

The data generating model is as follows:

$$
\begin{aligned}
Y_{t} &= 0.90 \cdot Y_{t-1} + 0.85 \cdot X_{t-1} + \epsilon_{X} \\
X_{t} &= 0.90 \cdot X_{t-1} + \epsilon_{Y} \\
\epsilon_{X}, \epsilon_{Y} &\sim \mathcal{N}(0, 0.05)
\end{aligned}
$$

We allow this coevolutionary process to play out over discrete timesteps. In 
each timestep, there is a small probability ($p$ = 0.05) of a speciation event, 
where the parent species splits into two child species. Coevolutionary dynamics 
then continue independently on each separate phylogenetic branch. We continue 
this process of evolution and speciation until we have a sufficient sample size
of species. We then fit our statistical model using the resulting phylogenetic
tree and the "contemporary" values of $X$ and $Y$ from the final timestep.

Note that this data generating process is a basic autoregressive model and, as 
such, is not identical to the statistical model. Despite this misspecification,
we find that, when fitted to data from 100 species, the statistical model 
converges normally and is able to accurately recover the true coevolutionary 
dynamics.

```{r echo=FALSE}
tar_load(deltaThetaX)
tar_load(deltaThetaY)
```

By plugging the fitted model parameters into Equation 3 above, we are able to 
calculate $\Delta\theta_{z}$, the change in the equilibrium value of each trait 
resulting from an absolute deviation increase in the other trait. Figure 
\@ref(fig:plotDT) visualises the resulting posterior distributions of 
$\Delta\theta_{z}$ for $X$ → $Y$ and $Y$ → $X$. We can see that the posterior 
distribution for $X$ → $Y$ is positive (median posterior value = 
`r print_num(median(deltaThetaY)[1,1])`, 
95% CI [`r print_num(quantile(deltaThetaY, 0.025)[1,1,1])` 
`r print_num(quantile(deltaThetaY, 0.975)[1,1,1])`]) suggesting that an absolute 
deviation increase in $X$ results in an increase in the equilibrium value of $Y$. 
However, the posterior distribution for $Y$ → $X$ includes zero (median 
posterior value = `r print_num(median(deltaThetaX)[1,1])`, 
95% CI [`r print_num(quantile(deltaThetaX, 0.025)[1,1,1])` 
`r print_num(quantile(deltaThetaX, 0.975)[1,1,1])`]) suggesting that an absolute 
deviation increase in $Y$ does not change the equilibrium value of $X$. This 
result accurately reflects the data generating process.

```{r plotDT, fig.cap="Posterior distributions for $\\Delta\\theta_{z}$ for both directions of coevolution. $\\Delta\\theta_{z}$ represents the expected z-score change in the equilibrium value of each trait resulting from an absolute deviation increase in the other trait.", fig.height=4, fig.width=4, echo=FALSE, warning=FALSE, message=FALSE}
tar_read(plotDT)
```

We can further visualise the influence of $X$ on $Y$ by plotting a phase plane 
of the evolutionary dynamics implied by the model. Figure \@ref(fig:plotPP) 
depicts the expected change in $X$ and $Y$ depending on their current states. 
Both variables have been standardised as z-scores for easier interpretation. 
The solid lines in the figure represent nullclines where $X$ (blue) and $Y$ 
(red) are at equilibrium, depending on the state of the other.

```{r plotPP, fig.cap="Phase plane showing the expected changes in $X$ and $Y$ depending on the state of the other trait. Arrows represent the direction and strength of selection in trait space. Solid lines represent nullclines where $X$ (blue) and $Y$ (red) are at equilibrium depending on the state of the other.", fig.height=6, fig.width=6, echo=FALSE, warning=FALSE, message=FALSE}
include_graphics("figures/phasePlane.pdf")
```

The phase plane shows that changes in the value of $X$ have large effects on 
the direction of selection on $Y$. When $X$ is low, there is strong negative 
selection on $Y$. When $X$ is high, there is strong positive selection on $Y$.
The nullclines also show that changes in $X$ have a large effect on the
equilibrium value for $Y$. However, changes in the value of $Y$ do not have the 
same effect on the direction of selection on $X$ or the equilibrium value of $X$.

To determine whether this pattern of results generalises beyond a single 
simulation run, we iterated the above simulation 100 times for three different 
sample sizes: 20, 50, and 100 species. Table \@ref(tab:power) shows the 
proportion of models where the posterior $\Delta\theta_{z}$ values have 95% 
credible intervals excluding zero. The results show that the model is relatively 
underpowered to detect the causal effect of $X$ on $Y$ when there are only 20 
species, but it is able to detect this causal effect with sufficient power when
the sample size is 50 or 100.

(ref:powerCaption) The proportion of models in which the 95% credible interval for $\Delta\theta_{z}$ is greater than zero for three different sample sizes. 100 simulations were run for each sample size. Four models did not converge ($\hat{R}$ values > 1.1) and were not included in these proportions.

```{r power, echo=FALSE, warning=FALSE, message=FALSE}
tar_read(power) %>%
  # do not show power for delta theta difference
  filter(Direction != "Difference") %>%
  # print table
  kable(
    format = "latex",
    digits = 2,
    booktabs = TRUE,
    caption = "(ref:powerCaption)"
    )
```

# 2.4.2. Evolution of social complexity across human societies

...

Sheehan et al. [-@Sheehan2023] studied the coevolution of religious and political 
authority in 97 Austronesian societies. Religious and political authority were 
both coded as four-level ordinal variables: absent, sub-local authority, local 
authority, and supra-local authority. Mapping these variables onto a phylogeny 
of Austronesian languages revealed that both religious and political authority 
had high phylogenetic signals, suggesting that a DPM could reasonably be used to 
assess the coevolution of these variables over the past 5000 years. Instead of 
binarising the two ordinal variables to use Pagel's discrete method, as previous
work had done [@Sheehan2018;@Watts2016], the authors explicitly modelled both 
variables as ordinal in the DPM. The model revealed that both religious and 
political authority coevolved reciprocally over time. In other words, increases
in religious authority led to strong positive selection on political authority
and, likewise, increases in political authority led to strong positive selection
on religious authority. This is akin to "runaway selection" where both traits
enter a positive feedback loop and increase concomitantly. This coevolutionary
relationship makes sense in light of Austronesian ethnographies, which describe
how both forms of authority are tightly interdependent and have have often
historically served to legitimise one another [_e.g._, @Goodenough2002].

# References

### 2024-06-24 R/calculatePower.R

@@ -10,5 +10,6 @@ calculatePower <- function(results) {
     pivot_wider(
       names_from = "n",
       values_from = "prop"
-    )
+    ) %>%
+    rename(Direction = direction)
 }

### 2024-06-24 R/packages.R

@@ -5,7 +5,8 @@ library(tarchetypes)
 library(tidyverse)
 tar_option_set(
   packages = c(
-    "ape","cmdstanr","coevolve","posterior","phaseR","tidybayes","tidyverse"
+    "ape","cmdstanr","coevolve","knitr","papaja",
+    "posterior","phaseR","tidybayes","tidyverse"
     )#,
   #controller = crew_controller_slurm(
   #  name = "coevolveSim",

### 2024-06-24 _targets.R

@@ -2,7 +2,7 @@ targets::tar_source()
 
 # pipeline
 list(
-  ### 1. Synthetic data simulation
+  ### Synthetic data simulation
   
   # compile model
   tar_target(model, compileModel()),
@@ -30,6 +30,11 @@ list(
   tar_target(plotDT, plotDeltaTheta(deltaThetaX, deltaThetaY)),
   # plot phase plane
   tar_target(plotPP, plotPhasePlane(post)),
+  
+  ### Report
+  
+  # render report
+  tar_render(report, "report.Rmd"),
   
   ### Session info
   tar_target(

### 2024-06-24 R/packages.R
Update README

@@ -6,7 +6,7 @@ library(tidyverse)
 tar_option_set(
   packages = c(
     "ape","cmdstanr","coevolve","knitr","papaja",
-    "posterior","phaseR","tidybayes","tidyverse"
+    "phaseR","posterior","tidybayes","tidyverse"
     )#,
   #controller = crew_controller_slurm(
   #  name = "coevolveSim",

### 2024-08-14 R/calculateDeltaTheta.R
Update workflow with new coevolve version

@@ -1,26 +1,0 @@
-calculateDeltaTheta <- function(draws, data, resp) {
-  # get variables
-  A <- draws$A
-  b <- draws$b
-  # medians and median absolute deviations for both variables
-  x <- draws$eta[1:data$N,1]
-  y <- draws$eta[1:data$N,2]
-  med_x <- rvar_median(x)
-  med_y <- rvar_median(y)
-  diff_x <- rvar_mad(x) 
-  diff_y <- rvar_mad(y)
-  # calculate delta theta
-  if (resp == "x") {
-    
-    med_theta <- -(med_y*A[1,2] + b[1]) / A[1,1]
-    diff_theta <- -((med_y + diff_y)*A[1,2] + b[1]) / A[1,1]
-    return( (diff_theta - med_theta) / diff_x )
-    
-  } else if (resp == "y") {
-    
-    med_theta <- -(med_x*A[2,1] + b[2]) / A[2,2]
-    diff_theta <- -((med_x + diff_x)*A[2,1] + b[2]) / A[2,2]
-    return( (diff_theta - med_theta) / diff_y )
-    
-  }
-}

### 2024-08-14 R/calculate_power.R

@@ -1,7 +1,7 @@
 calculatePower <- function(results) {
   results %>%
     # remove non-converging models
-    filter(maxRhat <= 1.1) %>%
+    filter(max_rhat <= 1.1) %>%
     # summarise proportions of effects with credible intervals above zero
     group_by(n, direction) %>%
     summarise(prop = mean(lower >= 0 & upper >= 0), .groups = "drop") %>%

### 2024-08-14 R/compile_model.R

@@ -1,4 +1,4 @@
-compileModel <- function() {
+compile_model <- function() {
   # generate dummy data and phylogenetic tree
   n <- 20
   tree <- ape::rcoal(n)

### 2024-08-14 R/fitModel.R

@@ -1,14 +1,0 @@
-fitModel <- function(model, data) {
-  # fit model
-  fit <- model$sample(
-    data = data,
-    chains = 4,
-    parallel_chains = 4,
-    output_dir = "out",
-    iter_warmup = 2000,
-    iter_sampling = 2000,
-    adapt_delta = 0.99,
-    seed = 1L
-    )
-  return(fit)
-}

### 2024-08-14 R/fitModelAndExtract.R

@@ -1,42 +1,0 @@
-fitModelAndExtract <- function(model, data, id) {
-  # fit model
-  fit <- model$sample(
-    data = data,
-    chains = 4,
-    parallel_chains = 4,
-    iter_warmup = 2000,
-    iter_sampling = 2000,
-    adapt_delta = 0.99,
-    seed = 1L
-    )
-  # get posterior
-  draws <- as_draws_rvars(fit)
-  # calculate delta theta for both variables
-  delta_theta_x <- calculateDeltaTheta(draws, data, resp = "x")
-  delta_theta_y <- calculateDeltaTheta(draws, data, resp = "y")
-  # save summary
-  tibble(
-    direction = c("X -> Y", "Y -> X", "Difference"),
-    delta_theta = list(delta_theta_y, delta_theta_x,
-                       delta_theta_y - delta_theta_x)
-    ) %>%
-    mutate(
-      median = map(delta_theta, function(x) as.numeric(median(x))),
-      lower  = map(delta_theta, function(x) as.numeric(quantile(x, 0.025))),
-      upper  = map(delta_theta, function(x) as.numeric(quantile(x, 0.975)))
-    ) %>%
-    dplyr::select(-delta_theta) %>%
-    unnest(c(median, lower, upper)) %>%
-    # index model number and sample size
-    mutate(
-      id = id, 
-      n = data$N
-      ) %>%
-    select(id, n, everything()) %>%
-    # add useful model information
-    mutate(
-      maxRhat = max(fit$summary()$rhat, na.rm = TRUE),
-      numDivergences = sum(fit$sampler_diagnostics()[,,"divergent__"]),
-      totalTimeSecs = fit$time()$total
-    )
-}

### 2024-08-14 R/fit_model.R

@@ -1,0 +1,42 @@
+fit_model <- function(model, sim, output_dir = getOption("cmdstanr_output_dir")) {
+  # get data for stan
+  sd <-
+    coevolve::coev_make_standata(
+      data = sim$data,
+      variables = list(
+        x = "normal",
+        y = "normal"
+      ),
+      id = "species",
+      tree = sim$tree
+    )
+  # fit model
+  fit <- 
+    model$sample(
+      data = sd,
+      chains = 4,
+      parallel_chains = 4,
+      output_dir = output_dir,
+      iter_warmup = 1000,
+      iter_sampling = 1000,
+      seed = 1L
+    )
+  # return object of class 'coevfit'
+  out <-
+    list(
+      fit = fit,
+      data = sim$data,
+      data_name = deparse(substitute(sim$data)),
+      variables = list(x = "normal", y = "normal"),
+      id = "species",
+      tree = sim$tree,
+      stan_code = model$code(),
+      stan_data = sd,
+      effects_mat = sd$effects_mat,
+      dist_mat = sd$dist_mat,
+      scale = TRUE,
+      prior_only = FALSE
+    )
+  class(out) <- "coevfit"
+  return(out)
+}

### 2024-08-14 R/fit_model_and_extract.R

@@ -1,0 +1,44 @@
+fit_model_and_extract <- function(model, sim, id = NULL) {
+  # fit model
+  fit <- fit_model(model, sim)
+  # calculate delta theta for both variables
+  delta_theta_x <- 
+    coevolve::coev_calculate_delta_theta(
+      object = fit,
+      response = "x",
+      predictor = "y"
+      )
+  delta_theta_y <- 
+    coevolve::coev_calculate_delta_theta(
+      object = fit,
+      response = "y",
+      predictor = "x"
+      )
+  # save summary
+  tibble(
+    direction = c(
+      "X -> Y",
+      "Y -> X"
+      ),
+    delta_theta = list(
+      as.numeric(delta_theta_y),
+      as.numeric(delta_theta_x)
+      )
+    ) %>%
+    mutate(
+      median = map(delta_theta, function(x) as.numeric(median(x))),
+      lower  = map(delta_theta, function(x) as.numeric(quantile(x, 0.025))),
+      upper  = map(delta_theta, function(x) as.numeric(quantile(x, 0.975)))
+    ) %>%
+    dplyr::select(-delta_theta) %>%
+    unnest(c(median, lower, upper)) %>%
+    # index model number and sample size
+    mutate(id = id, n = nrow(sim$data)) %>%
+    select(id, n, everything()) %>%
+    # add useful model information
+    mutate(
+      max_rhat = max(fit$fit$summary()$rhat, na.rm = TRUE),
+      num_divergences = sum(fit$fit$sampler_diagnostics()[,,"divergent__"]),
+      total_time_secs = fit$fit$time()$total
+    )
+}

### 2024-08-14 R/plotDeltaTheta.R

@@ -1,30 +1,0 @@
-plotDeltaTheta <- function(deltaThetaX, deltaThetaY) {
-  # get delta theta as vector
-  deltaThetaX <- as.numeric(as_draws_array(deltaThetaX))
-  deltaThetaY <- as.numeric(as_draws_array(deltaThetaY))
-  # plot posterior distribution
-  out <- 
-    tibble(
-      post = c(deltaThetaX, deltaThetaY),
-      Direction = rep(c("Y -> X", "X -> Y"), each = length(deltaThetaX))
-    ) %>%
-    ggplot(aes(x = post, fill = Direction)) +
-    stat_slab(alpha = 0.8) +
-    scale_x_continuous(
-      name = expression(paste(Delta, theta[z])),
-      limits = c(-10, 20),
-      expand = c(0, 0)
-      ) +
-    theme_classic() +
-    theme(
-      axis.title.y = element_blank(),
-      axis.ticks.y = element_blank(),
-      axis.line.y = element_blank(),
-      axis.text.y = element_blank(),
-      legend.position = c(0.9, 0.8),
-      plot.margin = margin(0, 5, 2, 5, "mm")
-    )
-  # save
-  ggsave(out, filename = "figures/deltaTheta.pdf", height = 4, width = 4)
-  return(out)
-}

### 2024-08-14 R/plotPhasePlane.R

@@ -1,90 +1,0 @@
-plotPhasePlane <- function(post) {
-  # get distribution of latent variables among species
-  median_X <- median(median(post$eta[1:100,1]))
-  median_Y <- median(median(post$eta[1:100,2]))
-  mad_X    <- mad(median(post$eta[1:100,1]))
-  mad_Y    <- mad(median(post$eta[1:100,2]))
-  low_X    <- median_X - mad_X*2.5 # -2.5 SD
-  high_X   <- median_X + mad_X*2.5 # +2.5 SD
-  low_Y    <- median_Y - mad_Y*2.5 # -2.5 SD
-  high_Y   <- median_Y + mad_Y*2.5 # +2.5 SD
-  # get parameter values
-  A <- median(post$A)
-  b <- median(post$b)
-  sigma <- median(post$sigma)
-  # function for flow field diagram
-  OU <- function(t, y, parameters) {
-    dy <- numeric(2)
-    dy[1] <- y[1]*A[1,1] + y[2]*A[1,2] + b[1]
-    dy[2] <- y[2]*A[2,2] + y[1]*A[2,1] + b[2]
-    list(dy)
-  }
-  # set up canvas
-  pdf(
-    paste0("figures/phasePlane.pdf"), 
-    width = 6,
-    height = 6,
-    pointsize = 12
-    )
-  par(pty = "s")
-  # create flow field diagram
-  OU.flowField <- 
-    flowField(
-      OU, 
-      xlim = c(low_X - 0.2, high_X + 0.2), 
-      ylim = c(low_Y - 0.2, high_Y + 0.2), 
-      parameters = NA,
-      add = FALSE, 
-      xlab = "",
-      ylab = "", 
-      points = 12,
-      col = "grey", 
-      xaxt = 'n',
-      yaxt = 'n', 
-      arrow.type = "proportional", 
-      frac = 1.5,
-      xaxs = "i", 
-      yaxs = "i", 
-      axes = FALSE,
-      lwd = 2
-    )
-  mtext(
-    side = 1,
-    "X (z-score)",
-    at = median_X,
-    line = 2.5,
-    cex = 1.3
-    )
-  mtext(
-    side = 2,
-    "Y (z-score)",
-    at = median_Y,
-    line = 2.5,
-    cex = 1.3
-    )
-  # add nullclines to phase plane
-  nc <- 
-    nullclines(
-      OU, 
-      xlim = c(low_X, high_X),
[…]

### 2024-08-14 R/plot_delta_theta.R

@@ -1,0 +1,44 @@
+plot_delta_theta <- function(fit) {
+  # get delta theta values
+  delta_theta_x <- 
+    as.numeric(
+      coevolve::coev_calculate_delta_theta(
+        object = fit,
+        response = "x",
+        predictor = "y"
+      )
+    )
+  delta_theta_y <- 
+    as.numeric(
+      coevolve::coev_calculate_delta_theta(
+        object = fit,
+        response = "y",
+        predictor = "x"
+      )
+    )
+  # plot posterior distribution
+  out <- 
+    tibble(
+      post = c(delta_theta_x, delta_theta_y),
+      Direction = rep(c("Y -> X", "X -> Y"), each = length(delta_theta_x))
+    ) %>%
+    ggplot(aes(x = post, fill = Direction)) +
+    stat_slab(alpha = 0.8) +
+    scale_x_continuous(
+      name = expression(paste(Delta, theta[z])),
+      limits = c(-10, 20),
+      expand = c(0, 0)
+    ) +
+    theme_classic() +
+    theme(
+      axis.title.y = element_blank(),
+      axis.ticks.y = element_blank(),
+      axis.line.y = element_blank(),
+      axis.text.y = element_blank(),
+      legend.position = c(0.9, 0.8),
+      plot.margin = margin(0, 5, 2, 5, "mm")
+    )
+  # save
+  ggsave(out, filename = "figures/delta_theta.pdf", height = 4, width = 4)
+  return(out)
+}

### 2024-08-14 R/plot_phase_plane.R

@@ -1,0 +1,16 @@
+plot_phase_plane <- function(fit) {
+  # set up canvas
+  pdf(
+    paste0("figures/phase_plane.pdf"), 
+    width = 6,
+    height = 6,
+    pointsize = 12
+  )
+  # plot flow field
+  coevolve::coev_plot_flowfield(
+    object = fit,
+    var1 = "x",
+    var2 = "y"
+  )
+  dev.off()
+}

### 2024-08-14 R/simulateData.R

@@ -1,31 +1,0 @@
-simulateData <- function(n, id) {
-  # set arguments for simulation
-  variables <- c("x","y")
-  selection_matrix <- 
-    matrix(
-      c(0.95, 0.00, # x -> y but not vice versa
-        0.95, 0.95),
-      nrow = 2,
-      ncol = 2,
-      byrow = TRUE,
-      dimnames = list(variables, variables)
-    )
-  drift <- c(
-    "x" = 0.05,
-    "y" = 0.05
-    )
-  # use function in coevolve package to simulate data
-  sim <- coevolve::coev_simulate_coevolution(
-    n, variables, selection_matrix, drift, prob_split = 0.05
-  )
-  # return data list for stan model
-  coevolve::coev_make_standata(
-    data = sim$data,
-    variables = list(
-      x = "normal",
-      y = "normal"
-    ),
-    id = "species",
-    tree = sim$tree
-  )
-}

### 2024-08-14 R/simulate_data.R

@@ -1,0 +1,21 @@
+simulate_data <- function(n, id = NULL) {
+  # set arguments for simulation
+  variables <- c("x","y")
+  selection_matrix <- 
+    matrix(
+      c(0.90, 0.00, # x -> y but not vice versa
+        0.85, 0.90),
+      nrow = 2,
+      ncol = 2,
+      byrow = TRUE,
+      dimnames = list(variables, variables)
+    )
+  drift <- c(
+    "x" = 0.05,
+    "y" = 0.05
+    )
+  # use function in coevolve package to return simulate data
+  coevolve::coev_simulate_coevolution(
+    n, variables, selection_matrix, drift, prob_split = 0.05
+  )
+}

### 2024-08-14 _targets.R

@@ -2,43 +2,44 @@ targets::tar_source()
 
 # pipeline
 list(
+  
   ### Synthetic data simulation
   
   # compile model
-  tar_target(model, compileModel()),
-  # simulate data and fit models
-  tar_target(id, 1:100),
+  tar_target(model, compile_model()),
+  # fit example individual model
+  tar_target(sim, simulate_data(n = 5)),
+  tar_target(fit, fit_model(model, sim, output_dir = "out")),
+  # plot figures
+  tar_target(phase_plane, plot_phase_plane(fit)),
+  tar_target(delta_theta, plot_delta_theta(fit)),
+  # simulate data and fit separate models
+  tar_target(id, 1:2),
   tar_map(
     # loop over different sample sizes
-    values = tibble(n = c(20, 50, 100)),
-    tar_target(data, simulateData(n, id), pattern = map(id)),
-    tar_target(results, fitModelAndExtract(model, data, id), 
-               pattern = map(data, id))
+    values = tibble(n = c(5, 10)),
+    tar_target(sim, simulate_data(n, id), pattern = map(id)),
+    tar_target(results, fit_model_and_extract(model, sim, id), 
+               pattern = map(sim, id))
   ),
   # combine results
-  tar_target(results, bind_rows(results_20, results_50, results_100)),
-  # summarise results
+  tar_target(results, bind_rows(results_5, results_10)),
+  # summarise power
   tar_target(power, calculatePower(results)),
-  # fit example individual model
-  tar_target(df, simulateData(n = 100, id = NULL)),
-  tar_target(fit, fitModel(model, df)),
-  tar_target(post, as_draws_rvars(fit)),
-  # calculate delta theta values
-  tar_target(deltaThetaX, calculateDeltaTheta(post, df, resp = "x")),
-  tar_target(deltaThetaY, calculateDeltaTheta(post, df, resp = "y")),
-  # plot delta theta
-  tar_target(plotDT, plotDeltaTheta(deltaThetaX, deltaThetaY)),
-  # plot phase plane
-  tar_target(plotPP, plotPhasePlane(post)),
   
-  ### Report
+  ### Simulation-based calibration
   
-  # render report
-  tar_render(report, "report.Rmd"),
+  ### Life history and niche complexity model
+  
+  ### Manuscript
+  
+  # render manuscript
+  #tar_render(manuscript, "manuscript.Rmd"),
   
   ### Session info
   tar_target(
[…]

### 2024-08-14 manuscript.Rmd
Add manuscript Rmd

---
title: "Causal inference and the detection of coevolutionary contingencies using dynamic phylogenetic models"
author: "E. Ringen, J. S. Martin, S. Claessens, & A. V. Jaeggi"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    number_sections: true
bibliography: references.bib
link-citations: yes
linkcolor: blue
urlcolor: black
figsintext : yes 
header-includes:
  \usepackage{fixltx2e}
  \usepackage{hyperref}
  \newcommand\figref[1]{{\hypersetup{linkcolor=blue}\autoref{#1}}}
---

\begin{center}
Abstract text here...
\end{center}

# Introduction

Phylogenetic comparative methods (PCMs) are commonly used to study the
co-evolution of organismal traits, spanning topics such as anatomy and 
physiology [@Dunn2015; @Garland2005; @Navalon2019; @OConnor2022; @Thayer2018], 
life history and behavior [@Bielby2007; @Clayton1994; @MacLean2012; 
@Salguero2016], and cultural evolution [@Mace2005; @Navarrete2016; @Watts2016]. 
"Coevolution" in its broadest sense refers to repeatable patterns of trait 
covariation over time, which can be investigated using a diverse family of 
statistical techniques depending on the research question and type of data 
available [@Garamszegi2014; @Harvey1991; @Nunn2011]. Many PCMs focus on 
understanding how traits coevolve by estimating evolutionary correlations and 
rates of change across phylogenetic trees, which can be used to better 
understand natural history and inform ancestral state reconstructions. However,
evolutionary correlation does not imply evolutionary causation between traits 
[@Shipley2016]. Thus, PCMs facilitating inference of why traits coevolve and 
change together across time are indispensable tools for testing 
macroevolutionary and ecological theory.

Among others, phylogenetic generalized linear (mixed) models [@Grafen1989; 
@Hadfield2010; @Symonds2014], phylogenetic path analysis [@GonzalezVoyer2014; 
@vonHardenberg2013], and Pagel's [-@Pagel1994] discrete method are the most 
popular approaches for assessing trait coevolution on macroevolutionary 
timescales. While each of these methods has clear benefits and performs well 
under specific scenarios, they are each limited in their generality by strong 
assumptions regarding the direction of causal effects among traits, the process
of evolutionary change, and/or the statistical properties of the traits under 
investigation. As we detail further below, these modeling assumptions are likely
to be violated in most datasets used for phylogenetic analysis, increasing the 
risk of inferential error and ultimately inhibiting our ability to explain the
dynamics of phenotypic evolution. We therefore introduce a novel class of PCMs
designed to address these challenges in a cohesive and flexible statistical
framework, using Bayesian MCMC algorithms in the Stan statistical programming
language [@Carpenter2017].

We begin by briefly reviewing the benefits and constraints of current PCMs, 
particularly with regard to causal inference, exploring how their limitations 
motivate further methods development. We then formally and conceptually 
introduce a novel class of dynamic phylogenetic models. We provide a worked 
synthetic example of coevolution between two continuous traits -- illustrated as
female promiscuity and sperm length [@Fitzpatrick2009] -- to assess the accuracy
and uncertainty of inferences made with dynamic phylogenetic models across a 
reasonable range of sample sizes for phylogenetic research. We also provide an 
accompanying coding tutorial using our **coevolve** R package to aid empiricists
in easily applying basic dynamic phylogenetic models to their own datasets. We 
then demonstrate the generality and flexibility of our method with two empirical
applications, which extend the model to complex, high-dimensional scenarios 
using dynamic latent variables. In particular, we show how our method can 
generate insights into the causal dynamics of trait coevolution across both 
genetic and cultural evolutionary timescales, using (1) a comparative dataset on
primate brain size, sociality, diet, and life history traits [@DeCasien2017], 
which reduced to two coevolving latent variables, and (2) two studies on the 
evolution of social complexity across a global and a regional sample of 
pre-industrial human societies [@Ringen_preprint; @Sheehan2023].

# Current approaches and motivation for a novel method

Fundamental to PCMs is the adjustment of raw trait associations for shared 
evolutionary history using a phylogenetic tree (or set of trees) and a 
statistical model. In a basic sense, phylogenetic adjustment is crucial for 
facilitating causal inference, as shared evolutionary history tends to generate 
trait correlations among closely related species with similar phenotypes, 
creating the illusion of convergent co-evolution even when traits evolve 
independently. Adjustment for phylogeny reduces bias and variance due to 
clustering over the tree, thus reducing the risks of type I (false-positive), 
type II (false-negative), type M (magnitude), and type S (sign) errors during 
statistical inference (*CITE*). Nonetheless, adjustment for phylogeny is not a 
magic fix for all sources of unobserved confounding, nor does it guarantee that 
resulting estimates are causally interpretable (*CITE*). Therefore, while all 
PCMs employ statistical techniques that reduce bias, these methods vary widely 
in the degree to which they isolate the causal pathways by which traits coevolve 
over time, with most commonly used methods focusing largely on evolutionary 
correlation (\figref{fig:dags}a) rather than causation (\figref{fig:dags}b-e).

```{=latex}
\begin{figure}
```

```{r dags, echo=FALSE, out.width='100%'}
knitr::include_graphics("figures/dags.png")
```

```{=latex}
\caption{Explanatory models of trait coevolution. \label{fig:dags}}
\textit{Note:} Examples of distinct formal approaches to describing and 
explaining patterns of coevolution among traits (bold letters), with important 
properties represented in each graph highlighted by blue arrows and text. 
Approaches range in complexity from (a) simple models of phylogenetic 
correlations (indicated by bidirectional arrows), which are useful for 
predictive purposes but fail to explain the causes of trait associations; to 
(b-e) directed acyclic graph models, which can be used to directly test the 
causal effects (directed arrows) driving trait associations across a phylogeny. 
As explained in Box 2 below, explicit causal models are crucial for deciding 
which traits should be included or excluded from a multivariate analysis to 
avoid potential biases due to phenomena such as so-called forks (b) and 
colliders (c). Causal models can also be further distinguished by whether they 
model relationships among traits as static (b-c) or dynamic (d-e) effects. Only 
dynamic models can be used to account for feedback processes 
(t\textsubscript{-1} → t → t\textsubscript{+1}) generated by reciprocal 
causation (blue arrows) among traits and autoregressive effects within traits 
(grey arrows) over time (d). For high dimensional problems, it is important to 
also consider whether inclusion of latent causes into the dynamic DAG (e), 
capturing dimensions of evolutionary integration among multiple traits, can 
provide a more parsimonious and theoretically insightful causal model (see 
\autoref{sec:empirical}).
\end{figure}
```

## Phylogenetic generalized linear (mixed) models

## Phylogenetic path analysis

## Pagel's discrete dynamic method

# Dynamic phylogenetic models

## Formal model

## Implementation

## Synthetic example

## Empirical applications

\label{sec:empirical}

### Brain size evolution in primates

### Evolution of social complexity across human societies

# Conclusion

# References

### 2024-08-14 R/fit_model.R

@@ -17,8 +17,8 @@ fit_model <- function(model, sim, output_dir = getOption("cmdstanr_output_dir"))
       chains = 4,
       parallel_chains = 4,
       output_dir = output_dir,
-      iter_warmup = 1000,
-      iter_sampling = 1000,
+      iter_warmup = 2000,
+      iter_sampling = 2000,
       seed = 1L
     )
   # return object of class 'coevfit'

### 2024-08-14 _targets.R

@@ -8,7 +8,7 @@ list(
   # compile model
   tar_target(model, compile_model()),
   # fit example individual model
-  tar_target(sim, simulate_data(n = 5)),
+  tar_target(sim, simulate_data(n = 100)),
   tar_target(fit, fit_model(model, sim, output_dir = "out")),
   # plot figures
   tar_target(phase_plane, plot_phase_plane(fit)),
@@ -34,7 +34,7 @@ list(
   ### Manuscript
   
   # render manuscript
-  #tar_render(manuscript, "manuscript.Rmd"),
+  tar_render(manuscript, "manuscript.Rmd"),
   
   ### Session info
   tar_target(

### 2024-08-19 R/compile_model.R
Continue writing manuscript Rmd

@@ -2,27 +2,32 @@ compile_model <- function() {
   # generate dummy data and phylogenetic tree
   n <- 20
   tree <- ape::rcoal(n)
-  d <- data.frame(
-    id = tree$tip.label,
-    x = rnorm(n),
-    y = rnorm(n)
-  )
-  # make stan code 
-  scode <- coevolve::coev_make_stancode(
-    data = d,
-    variables = list(
-      x = "normal",
-      y = "normal"
-    ),
-    id = "id",
-    tree = tree
-  )
-  # write stan code to file
-  file <- cmdstanr::write_stan_file(
-    scode,
-    dir = "stan",
-    basename = "model"
+  d <- 
+    data.frame(
+      id = tree$tip.label,
+      Promiscuity = rnorm(n),
+      SpermSize   = rnorm(n)
     )
+  # make stan code 
+  scode <- 
+    coevolve::coev_make_stancode(
+      data = d,
+      variables = list(
+        Promiscuity = "normal",
+        SpermSize   = "normal"
+      ),
+      id = "id",
+      tree = tree
+    )
+  # write stan code to file
+  file <- 
+    cmdstanr::write_stan_file(
+      scode,
+      dir = "stan",
+      basename = "model"
+      )
   # compile model
-  cmdstanr::cmdstan_model(stan_file = file)
+  cmdstanr::cmdstan_model(
+    stan_file = file
+    )
 }

### 2024-08-19 R/fit_model.R

@@ -4,8 +4,8 @@ fit_model <- function(model, sim, output_dir = getOption("cmdstanr_output_dir"))
     coevolve::coev_make_standata(
       data = sim$data,
       variables = list(
-        x = "normal",
-        y = "normal"
+        Promiscuity = "normal",
+        SpermSize   = "normal"
       ),
       id = "species",
       tree = sim$tree
@@ -27,7 +27,10 @@ fit_model <- function(model, sim, output_dir = getOption("cmdstanr_output_dir"))
       fit = fit,
       data = sim$data,
       data_name = deparse(substitute(sim$data)),
-      variables = list(x = "normal", y = "normal"),
+      variables = list(
+        Promiscuity = "normal",
+        SpermSize   = "normal"
+        ),
       id = "species",
       tree = sim$tree,
       stan_code = model$code(),

### 2024-08-19 R/fit_model_and_extract.R

@@ -2,27 +2,27 @@ fit_model_and_extract <- function(model, sim, id = NULL) {
   # fit model
   fit <- fit_model(model, sim)
   # calculate delta theta for both variables
-  delta_theta_x <- 
+  delta_theta_promiscuity <- 
     coevolve::coev_calculate_delta_theta(
       object = fit,
-      response = "x",
-      predictor = "y"
+      response = "Promiscuity",
+      predictor = "SpermSize"
       )
-  delta_theta_y <- 
+  delta_theta_spermsize <- 
     coevolve::coev_calculate_delta_theta(
       object = fit,
-      response = "y",
-      predictor = "x"
+      response = "SpermSize",
+      predictor = "Promiscuity"
       )
   # save summary
   tibble(
     direction = c(
-      "X -> Y",
-      "Y -> X"
+      "Promiscuity -> Sperm size",
+      "Sperm size -> Promiscuity"
       ),
     delta_theta = list(
-      as.numeric(delta_theta_y),
-      as.numeric(delta_theta_x)
+      as.numeric(delta_theta_spermsize),
+      as.numeric(delta_theta_promiscuity)
       )
     ) %>%
     mutate(

### 2024-08-19 R/packages.R

@@ -5,7 +5,7 @@ library(tarchetypes)
 library(tidyverse)
 tar_option_set(
   packages = c(
-    "ape","cmdstanr","coevolve","knitr","papaja",
+    "ape","bookdown","cmdstanr","coevolve","cowplot","knitr",
     "phaseR","posterior","tidybayes","tidyverse"
     )#,
   #controller = crew_controller_slurm(

### 2024-08-19 R/plot_delta_theta.R

@@ -1,44 +1,0 @@
-plot_delta_theta <- function(fit) {
-  # get delta theta values
-  delta_theta_x <- 
-    as.numeric(
-      coevolve::coev_calculate_delta_theta(
-        object = fit,
-        response = "x",
-        predictor = "y"
-      )
-    )
-  delta_theta_y <- 
-    as.numeric(
-      coevolve::coev_calculate_delta_theta(
-        object = fit,
-        response = "y",
-        predictor = "x"
-      )
-    )
-  # plot posterior distribution
-  out <- 
-    tibble(
-      post = c(delta_theta_x, delta_theta_y),
-      Direction = rep(c("Y -> X", "X -> Y"), each = length(delta_theta_x))
-    ) %>%
-    ggplot(aes(x = post, fill = Direction)) +
-    stat_slab(alpha = 0.8) +
-    scale_x_continuous(
-      name = expression(paste(Delta, theta[z])),
-      limits = c(-10, 20),
-      expand = c(0, 0)
-    ) +
-    theme_classic() +
-    theme(
-      axis.title.y = element_blank(),
-      axis.ticks.y = element_blank(),
-      axis.line.y = element_blank(),
-      axis.text.y = element_blank(),
-      legend.position = c(0.9, 0.8),
-      plot.margin = margin(0, 5, 2, 5, "mm")
-    )
-  # save
-  ggsave(out, filename = "figures/delta_theta.pdf", height = 4, width = 4)
-  return(out)
-}

### 2024-08-19 R/plot_phase_plane.R

@@ -1,16 +1,0 @@
-plot_phase_plane <- function(fit) {
-  # set up canvas
-  pdf(
-    paste0("figures/phase_plane.pdf"), 
-    width = 6,
-    height = 6,
-    pointsize = 12
-  )
-  # plot flow field
-  coevolve::coev_plot_flowfield(
-    object = fit,
-    var1 = "x",
-    var2 = "y"
-  )
-  dev.off()
-}

### 2024-08-19 R/plot_synthetic_example.R

@@ -1,0 +1,102 @@
+plot_synthetic_example <- function(fit, delta_theta_promiscuity,
+                                   delta_theta_spermsize) {
+  # plot cichlid species
+  pA <-
+    ggdraw() +
+    draw_image(
+      image = "figures/cichlids/cichlids.png",
+      scale = 0.95
+      )
+  # plot delta theta
+  pB <- 
+    tibble(
+      post = c(
+        as.numeric(delta_theta_promiscuity),
+        as.numeric(delta_theta_spermsize)
+        ),
+      Direction = rep(
+        c("Sperm size -> Promiscuity",
+          "Promiscuity -> Sperm size"), 
+        each = length(delta_theta_promiscuity)
+        )
+    ) %>%
+    ggplot(aes(x = post, fill = Direction)) +
+    stat_slab(alpha = 0.8) +
+    scale_x_continuous(
+      name = expression(paste(Delta, theta[z])),
+      limits = c(-5, 15),
+      expand = c(0, 0)
+    ) +
+    scale_fill_manual(values = c("#c55852", "#5387b6")) +
+    theme_classic() +
+    theme(
+      axis.title.y = element_blank(),
+      axis.ticks.y = element_blank(),
+      axis.line.y = element_blank(),
+      axis.text.y = element_blank(),
+      legend.position = c(0.72, 0.8),
+      plot.margin = margin(0, 5, 2, 5, "mm"),
+      legend.text = element_text(size = 7),
+      legend.title = element_text(size = 7),
+      legend.key.size = unit(3, "mm")
+    )
+  # plot phase plane
+  png(
+    paste0("figures/cichlids/phase_plane.png"),
+    height = 12,
+    width = 12,
+    units = "cm",
+    res = 1200
+    )
+  coevolve::coev_plot_flowfield(
+    object = fit,
+    var1 = "Promiscuity",
+    var2 = "SpermSize",
+    nullclines = TRUE
+  )
+  dev.off()
+  pC <-
+    ggdraw() + 
+      draw_image(
+        image = "figures/cichlids/phase_plane.png"
+        ) +
+      draw_image(
+        image = "figures/cichlids/low_sperm_size.png",
+        scale = 0.15,
+        x = -0.40,
+        y = -0.27
+        ) +
+      draw_image(
+        image = "figures/cichlids/high_sperm_size.png",
+        scale = 0.25,
+        x = -0.43,
+        y =  0.31
[…]

### 2024-08-19 R/simulate_data.R

@@ -1,9 +1,9 @@
 simulate_data <- function(n, id = NULL) {
   # set arguments for simulation
-  variables <- c("x","y")
+  variables <- c("Promiscuity", "SpermSize")
   selection_matrix <- 
     matrix(
-      c(0.90, 0.00, # x -> y but not vice versa
+      c(0.90, 0.00, # promiscuity -> spermsize but not vice versa
         0.85, 0.90),
       nrow = 2,
       ncol = 2,
@@ -11,8 +11,8 @@ simulate_data <- function(n, id = NULL) {
       dimnames = list(variables, variables)
     )
   drift <- c(
-    "x" = 0.05,
-    "y" = 0.05
+    "Promiscuity" = 0.05,
+    "SpermSize"   = 0.05
     )
   # use function in coevolve package to return simulate data
   coevolve::coev_simulate_coevolution(

### 2024-08-19 _targets.R

@@ -10,9 +10,29 @@ list(
   # fit example individual model
   tar_target(sim, simulate_data(n = 100)),
   tar_target(fit, fit_model(model, sim, output_dir = "out")),
-  # plot figures
-  tar_target(phase_plane, plot_phase_plane(fit)),
-  tar_target(delta_theta, plot_delta_theta(fit)),
+  # get delta theta values
+  tar_target(
+    delta_theta_promiscuity,
+    coevolve::coev_calculate_delta_theta(
+      object = fit,
+      response = "Promiscuity",
+      predictor = "SpermSize"
+    )
+  ),
+  tar_target(
+    delta_theta_spermsize,
+    coevolve::coev_calculate_delta_theta(
+      object = fit,
+      response = "SpermSize",
+      predictor = "Promiscuity"
+    )
+  ),
+  # plot synthetic example
+  tar_target(
+    plot_synthetic,
+    plot_synthetic_example(fit, delta_theta_promiscuity, 
+                           delta_theta_spermsize)
+    ),
   # simulate data and fit separate models
   tar_target(id, 1:2),
   tar_map(

### 2024-08-20 R/bind_sbc_results.R
Add simulation-based calibration

@@ -1,0 +1,8 @@
+bind_sbc_results <- function(sbc_results) {
+  # loop over list and bind results
+  out <- sbc_results[[1]]
+  for (i in 2:length(sbc_results)) {
+    out <- SBC::bind_results(out, sbc_results[[i]])
+  }
+  return(out)
+}

### 2024-08-20 R/packages.R

@@ -6,7 +6,7 @@ library(tidyverse)
 tar_option_set(
   packages = c(
     "ape","bookdown","cmdstanr","coevolve","cowplot","knitr",
-    "phaseR","posterior","tidybayes","tidyverse"
+    "phaseR","posterior","SBC","tidybayes","tidyverse"
     )#,
   #controller = crew_controller_slurm(
   #  name = "coevolveSim",

### 2024-08-20 R/plot_sbc_ecdf.R

@@ -1,0 +1,17 @@
+plot_sbc_ecdf <- function(results_coev) {
+  # plot ECDF
+  out <-
+    SBC::plot_ecdf(results_coev) + 
+    ylab("ECDF") + 
+    xlab("Fractional rank") +
+    theme_classic() + 
+    theme(strip.background = element_blank())
+  # save
+  ggsave(
+    filename = "figures/simulation_based_calibration/ecdf.pdf",
+    plot = out,
+    width = 7,
+    height = 5
+  )
+  return(out)
+}

### 2024-08-20 R/plot_sbc_ecdf_diff.R

@@ -1,0 +1,18 @@
+plot_sbc_ecdf_diff <- function(results_coev) {
+  # plot ECDF difference
+  out <-
+    SBC::plot_ecdf_diff(results_coev) + 
+    scale_y_continuous(limits = c(-0.17, 0.17)) + 
+    ylab("ECDF difference") + 
+    xlab("Fractional rank") + 
+    theme_classic() + 
+    theme(strip.background = element_blank())
+  # save
+  ggsave(
+    filename = "figures/simulation_based_calibration/ecdf_difference.pdf",
+    plot = out,
+    width = 7,
+    height = 5
+  )
+  return(out)
+}

### 2024-08-20 R/prepare_sbc_backend.R

@@ -1,0 +1,11 @@
+prepare_sbc_backend <- function(sbc_model) {
+  # prepare backend
+  SBC::SBC_backend_cmdstan_sample(
+    model = sbc_model,
+    iter_warmup = 250,
+    iter_sampling = 1000,
+    chains = 2,
+    adapt_delta = 0.9,
+    seed = 1
+  )
+}

### 2024-08-20 R/prepare_sbc_generator.R

@@ -1,0 +1,8 @@
+prepare_sbc_generator <- function(sbc_model) {
+  # prepare generator
+  SBC::SBC_generator_function(
+    f = sbc_generator,
+    model = sbc_model,
+    N = 20
+  )
+}

### 2024-08-20 R/sbc_generator.R

@@ -1,0 +1,67 @@
+sbc_generator <- function(model, N) {
+  # simulate tree
+  tree <- ape::rcoal(N)
+  # simulate data
+  d_temp <- data.frame(
+    x = rnorm(N, 0, 1),
+    y = rbinom(N, 1, 0.5),
+    id = tree$tip.label
+  )
+  # get data list for stan
+  standat_temp <- 
+    coevolve::coev_make_standata(
+      data = d_temp,
+      variables = list(
+        x = "normal",
+        y = "bernoulli_logit"
+      ),
+      id = "id",
+      tree = tree,
+      prior_only = TRUE
+    )
+  # number of samples
+  n_samps <- 50
+  # fit model
+  sims <- model$sample(
+    data = standat_temp,
+    chains = 1,
+    iter_sampling = n_samps,
+    iter_warmup = n_samps
+  )
+  # get draws
+  draws <- posterior::as_draws_rvars(sims$draws())
+  A <- posterior::draws_of(draws$A)
+  Q <- posterior::draws_of(draws$Q)
+  b <- posterior::draws_of(draws$b)
+  log_likelihood <- posterior::draws_of(draws$log_likelihood)
+  x_sim <- posterior::draws_of(draws$yrep)[n_samps,,1]
+  y_sim <- posterior::draws_of(draws$yrep)[n_samps,,2]
+  # output list
+  list(
+    variables = list(
+      "A[1,1]" = A[n_samps,1,1],
+      "A[1,2]" = A[n_samps,1,2],
+      "A[2,2]" = A[n_samps,2,2],
+      "A[2,1]" = A[n_samps,2,1],
+      "b[1]"   = b[n_samps,1],
+      "b[2]"   = b[n_samps,2],
+      "Q[1,1]" = Q[n_samps,1,1],
+      "Q[2,2]" = Q[n_samps,2,2],
+      "log_likelihood" = log_likelihood[n_samps]
+    ),
+    generated = coev_make_standata(
+      data = data.frame(
+        x = x_sim,
+        y = as.integer(y_sim),
+        id = tree$tip.label
+        ),
+      variables = list(
+        x = "normal",
+        y = "bernoulli_logit"
+      ),
+      id = "id",
+      tree = tree,
+      prior_only = FALSE
+    )
+  )
+}

### 2024-08-20 _targets.R

@@ -49,6 +49,35 @@ list(
   
   ### Simulation-based calibration
   
+  # model file
+  tar_target(sbc_model_file, "stan/sbc_model.stan", format = "file"),
+  # compile model
+  tar_target(sbc_model, cmdstanr::cmdstan_model(sbc_model_file)),
+  # prepare backend and generator
+  tar_target(sbc_backend, prepare_sbc_backend(sbc_model)),
+  tar_target(sbc_generator, prepare_sbc_generator(sbc_model)),
+  # iterate over sbc_id
+  tar_target(sbc_id, 1:2),
+  # generate datasets
+  tar_target(
+    sbc_datasets,
+    SBC::generate_datasets(sbc_generator, n_sims = 1),
+    pattern = map(id),
+    iteration = "list"
+    ),
+  # compute sbc
+  tar_target(
+    sbc_results,
+    SBC::compute_SBC(sbc_datasets, sbc_backend),
+    pattern = map(sbc_datasets, id),
+    iteration = "list"
+    ),
+  # combine sbc results
+  tar_target(sbc_results_combined, bind_sbc_results(sbc_results)),
+  # plot sbc results
+  tar_target(plot_sbc1, plot_sbc_ecdf(sbc_results_combined)),
+  tar_target(plot_sbc2, plot_sbc_ecdf_diff(sbc_results_combined)),
+  
   ### Life history and niche complexity model
   
   ### Manuscript

### 2024-08-23 R/fit_model.R
Update manuscript with full simulations

@@ -19,6 +19,7 @@ fit_model <- function(model, sim, output_dir = getOption("cmdstanr_output_dir"))
       output_dir = output_dir,
       iter_warmup = 2000,
       iter_sampling = 2000,
+      adapt_delta = 0.99,
       seed = 1L
     )
   # return object of class 'coevfit'

### 2024-08-23 R/packages.R

@@ -9,11 +9,11 @@ tar_option_set(
     "phaseR","posterior","SBC","tidybayes","tidyverse"
     )#,
   #controller = crew_controller_slurm(
-  #  name = "coevolveSim",
-  #  workers = 1,
+  #  name = "dpm",
+  #  workers = 80,
   #  script_lines = "module load R",
-  #  slurm_time_minutes = 2880,
-  #  slurm_memory_gigabytes_per_cpu = 1,
+  #  slurm_time_minutes = 1440*2,
+  #  slurm_memory_gigabytes_per_cpu = 2,
   #  slurm_cpus_per_task = 4
   #)
 )

### 2024-08-23 R/prepare_sbc_generator.R

@@ -1,7 +1,7 @@
 prepare_sbc_generator <- function(sbc_model) {
   # prepare generator
   SBC::SBC_generator_function(
-    f = sbc_generator,
+    f = sbc_generator_function,
     model = sbc_model,
     N = 20
   )

### 2024-08-23 R/sbc_generator.R

@@ -1,67 +1,0 @@
-sbc_generator <- function(model, N) {
-  # simulate tree
-  tree <- ape::rcoal(N)
-  # simulate data
-  d_temp <- data.frame(
-    x = rnorm(N, 0, 1),
-    y = rbinom(N, 1, 0.5),
-    id = tree$tip.label
-  )
-  # get data list for stan
-  standat_temp <- 
-    coevolve::coev_make_standata(
-      data = d_temp,
-      variables = list(
-        x = "normal",
-        y = "bernoulli_logit"
-      ),
-      id = "id",
-      tree = tree,
-      prior_only = TRUE
-    )
-  # number of samples
-  n_samps <- 50
-  # fit model
-  sims <- model$sample(
-    data = standat_temp,
-    chains = 1,
-    iter_sampling = n_samps,
-    iter_warmup = n_samps
-  )
-  # get draws
-  draws <- posterior::as_draws_rvars(sims$draws())
-  A <- posterior::draws_of(draws$A)
-  Q <- posterior::draws_of(draws$Q)
-  b <- posterior::draws_of(draws$b)
-  log_likelihood <- posterior::draws_of(draws$log_likelihood)
-  x_sim <- posterior::draws_of(draws$yrep)[n_samps,,1]
-  y_sim <- posterior::draws_of(draws$yrep)[n_samps,,2]
-  # output list
-  list(
-    variables = list(
-      "A[1,1]" = A[n_samps,1,1],
-      "A[1,2]" = A[n_samps,1,2],
-      "A[2,2]" = A[n_samps,2,2],
-      "A[2,1]" = A[n_samps,2,1],
-      "b[1]"   = b[n_samps,1],
-      "b[2]"   = b[n_samps,2],
-      "Q[1,1]" = Q[n_samps,1,1],
-      "Q[2,2]" = Q[n_samps,2,2],
-      "log_likelihood" = log_likelihood[n_samps]
-    ),
-    generated = coev_make_standata(
-      data = data.frame(
-        x = x_sim,
-        y = as.integer(y_sim),
-        id = tree$tip.label
-        ),
-      variables = list(
-        x = "normal",
-        y = "bernoulli_logit"
-      ),
-      id = "id",
-      tree = tree,
-      prior_only = FALSE
-    )
-  )
-}

### 2024-08-23 R/sbc_generator_function.R

@@ -1,0 +1,67 @@
+sbc_generator_function <- function(model, N) {
+  # simulate tree
+  tree <- ape::rcoal(N)
+  # simulate data
+  d_temp <- data.frame(
+    x = rnorm(N, 0, 1),
+    y = rbinom(N, 1, 0.5),
+    id = tree$tip.label
+  )
+  # get data list for stan
+  standat_temp <- 
+    coevolve::coev_make_standata(
+      data = d_temp,
+      variables = list(
+        x = "normal",
+        y = "bernoulli_logit"
+      ),
+      id = "id",
+      tree = tree,
+      prior_only = TRUE
+    )
+  # number of samples
+  n_samps <- 50
+  # fit model
+  sims <- model$sample(
+    data = standat_temp,
+    chains = 1,
+    iter_sampling = n_samps,
+    iter_warmup = n_samps
+  )
+  # get draws
+  draws <- posterior::as_draws_rvars(sims$draws())
+  A <- posterior::draws_of(draws$A)
+  Q <- posterior::draws_of(draws$Q)
+  b <- posterior::draws_of(draws$b)
+  log_likelihood <- posterior::draws_of(draws$log_likelihood)
+  x_sim <- posterior::draws_of(draws$yrep)[n_samps,,1]
+  y_sim <- posterior::draws_of(draws$yrep)[n_samps,,2]
+  # output list
+  list(
+    variables = list(
+      "A[1,1]" = A[n_samps,1,1],
+      "A[1,2]" = A[n_samps,1,2],
+      "A[2,2]" = A[n_samps,2,2],
+      "A[2,1]" = A[n_samps,2,1],
+      "b[1]"   = b[n_samps,1],
+      "b[2]"   = b[n_samps,2],
+      "Q[1,1]" = Q[n_samps,1,1],
+      "Q[2,2]" = Q[n_samps,2,2],
+      "log_likelihood" = log_likelihood[n_samps]
+    ),
+    generated = coev_make_standata(
+      data = data.frame(
+        x = x_sim,
+        y = as.integer(y_sim),
+        id = tree$tip.label
+        ),
+      variables = list(
+        x = "normal",
+        y = "bernoulli_logit"
+      ),
+      id = "id",
+      tree = tree,
+      prior_only = FALSE
+    )
+  )
+}

### 2024-08-23 _targets.R

@@ -34,16 +34,16 @@ list(
                            delta_theta_spermsize)
     ),
   # simulate data and fit separate models
-  tar_target(id, 1:2),
+  tar_target(id, 1:100),
   tar_map(
     # loop over different sample sizes
-    values = tibble(n = c(5, 10)),
+    values = tibble(n = c(20, 50, 100)),
     tar_target(sim, simulate_data(n, id), pattern = map(id)),
     tar_target(results, fit_model_and_extract(model, sim, id), 
                pattern = map(sim, id))
   ),
   # combine results
-  tar_target(results, bind_rows(results_5, results_10)),
+  tar_target(results, bind_rows(results_20, results_50, results_100)),
   # summarise power
   tar_target(power, calculatePower(results)),
   
@@ -56,20 +56,20 @@ list(
   # prepare backend and generator
   tar_target(sbc_backend, prepare_sbc_backend(sbc_model)),
   tar_target(sbc_generator, prepare_sbc_generator(sbc_model)),
-  # iterate over sbc_id
-  tar_target(sbc_id, 1:2),
+  # iterate over sbc_id batches
+  tar_target(sbc_id, 1:50),
   # generate datasets
   tar_target(
     sbc_datasets,
-    SBC::generate_datasets(sbc_generator, n_sims = 1),
-    pattern = map(id),
+    SBC::generate_datasets(sbc_generator, n_sims = 10),
+    pattern = map(sbc_id),
     iteration = "list"
     ),
   # compute sbc
   tar_target(
     sbc_results,
     SBC::compute_SBC(sbc_datasets, sbc_backend),
-    pattern = map(sbc_datasets, id),
+    pattern = map(sbc_datasets, sbc_id),
     iteration = "list"
     ),
   # combine sbc results

### 2024-08-23 R/packages.R
Remove seed in SBC

@@ -10,9 +10,9 @@ tar_option_set(
     )#,
   #controller = crew_controller_slurm(
   #  name = "dpm",
-  #  workers = 80,
+  #  workers = 50,
   #  script_lines = "module load R",
-  #  slurm_time_minutes = 1440*2,
+  #  slurm_time_minutes = 60,
   #  slurm_memory_gigabytes_per_cpu = 2,
   #  slurm_cpus_per_task = 4
   #)

### 2024-08-23 R/prepare_sbc_backend.R

@@ -5,7 +5,6 @@ prepare_sbc_backend <- function(sbc_model) {
     iter_warmup = 250,
     iter_sampling = 1000,
     chains = 2,
-    adapt_delta = 0.9,
-    seed = 1
+    adapt_delta = 0.9
   )
 }

### 2024-08-23 R/packages.R
Add scale = FALSE argument to SBC

@@ -12,7 +12,7 @@ tar_option_set(
   #  name = "dpm",
   #  workers = 50,
   #  script_lines = "module load R",
-  #  slurm_time_minutes = 60,
+  #  slurm_time_minutes = 20,
   #  slurm_memory_gigabytes_per_cpu = 2,
   #  slurm_cpus_per_task = 4
   #)

### 2024-08-23 R/sbc_generator_function.R

@@ -17,6 +17,7 @@ sbc_generator_function <- function(model, N) {
       ),
       id = "id",
       tree = tree,
+      scale = FALSE,
       prior_only = TRUE
     )
   # number of samples
@@ -61,6 +62,7 @@ sbc_generator_function <- function(model, N) {
       ),
       id = "id",
       tree = tree,
+      scale = FALSE,
       prior_only = FALSE
     )
   )

### 2024-08-23 _targets.R

@@ -73,7 +73,8 @@ list(
     iteration = "list"
     ),
   # combine sbc results
-  tar_target(sbc_results_combined, bind_sbc_results(sbc_results)),
+  tar_target(sbc_results_combined, bind_sbc_results(sbc_results),
+             deployment = "main"),
   # plot sbc results
   tar_target(plot_sbc1, plot_sbc_ecdf(sbc_results_combined)),
   tar_target(plot_sbc2, plot_sbc_ecdf_diff(sbc_results_combined)),

### 2024-08-23 R/packages.R
Use here package for file paths

@@ -5,8 +5,8 @@ library(tarchetypes)
 library(tidyverse)
 tar_option_set(
   packages = c(
-    "ape","bookdown","cmdstanr","coevolve","cowplot","knitr",
-    "phaseR","posterior","SBC","tidybayes","tidyverse"
+    "ape","bookdown","cmdstanr","coevolve","cowplot","here",
+    "knitr","phaseR","posterior","SBC","tidybayes","tidyverse"
     )#,
   #controller = crew_controller_slurm(
   #  name = "dpm",

### 2024-08-23 R/plot_sbc_ecdf.R

@@ -8,7 +8,7 @@ plot_sbc_ecdf <- function(results_coev) {
     theme(strip.background = element_blank())
   # save
   ggsave(
-    filename = "figures/simulation_based_calibration/ecdf.pdf",
+    filename = here::here("figures/simulation_based_calibration/ecdf.pdf"),
     plot = out,
     width = 7,
     height = 5

### 2024-08-23 R/plot_sbc_ecdf_diff.R

@@ -9,7 +9,7 @@ plot_sbc_ecdf_diff <- function(results_coev) {
     theme(strip.background = element_blank())
   # save
   ggsave(
-    filename = "figures/simulation_based_calibration/ecdf_difference.pdf",
+    filename = here::here("figures/simulation_based_calibration/ecdf_difference.pdf"),
     plot = out,
     width = 7,
     height = 5

### 2024-08-23 R/plot_synthetic_example.R

@@ -4,7 +4,7 @@ plot_synthetic_example <- function(fit, delta_theta_promiscuity,
   pA <-
     ggdraw() +
     draw_image(
-      image = "figures/cichlids/cichlids.png",
+      image = here::here("figures/cichlids/cichlids.png"),
       scale = 0.95
       )
   # plot delta theta
@@ -42,7 +42,7 @@ plot_synthetic_example <- function(fit, delta_theta_promiscuity,
     )
   # plot phase plane
   png(
-    paste0("figures/cichlids/phase_plane.png"),
+    here::here("figures/cichlids/phase_plane.png"),
     height = 12,
     width = 12,
     units = "cm",
@@ -58,28 +58,28 @@ plot_synthetic_example <- function(fit, delta_theta_promiscuity,
   pC <-
     ggdraw() + 
       draw_image(
-        image = "figures/cichlids/phase_plane.png"
+        image = here::here("figures/cichlids/phase_plane.png")
         ) +
       draw_image(
-        image = "figures/cichlids/low_sperm_size.png",
+        image = here::here("figures/cichlids/low_sperm_size.png"),
         scale = 0.15,
         x = -0.40,
         y = -0.27
         ) +
       draw_image(
-        image = "figures/cichlids/high_sperm_size.png",
+        image = here::here("figures/cichlids/high_sperm_size.png"),
         scale = 0.25,
         x = -0.43,
         y =  0.31
       ) +
       draw_image(
-        image = "figures/cichlids/low_promiscuity.png",
+        image = here::here("figures/cichlids/low_promiscuity.png"),
         scale = 0.15,
         x = -0.25,
         y = -0.41
       ) +
       draw_image(
-        image = "figures/cichlids/high_promiscuity.png",
+        image = here::here("figures/cichlids/high_promiscuity.png"),
         scale = 0.15,
         x =  0.35,
         y = -0.41
@@ -94,7 +94,7 @@ plot_synthetic_example <- function(fit, delta_theta_promiscuity,
     )
   ggsave(
     plot = out,
-    filename = "figures/cichlids/synthetic.pdf",
+    filename = here::here("figures/cichlids/synthetic.pdf"),
     width = 7,
     height = 3
     )

### 2024-08-27 R/packages.R
Fix cowplot issues

@@ -5,8 +5,8 @@ library(tarchetypes)
 library(tidyverse)
 tar_option_set(
   packages = c(
-    "ape","bookdown","cmdstanr","coevolve","cowplot","here",
-    "knitr","phaseR","posterior","SBC","tidybayes","tidyverse"
+    "ape","bookdown","cmdstanr","coevolve","cowplot","knitr",
+    "magick","phaseR","posterior","SBC","tidybayes","tidyverse"
     )#,
   #controller = crew_controller_slurm(
   #  name = "dpm",

### 2024-08-27 R/plot_sbc_ecdf.R

@@ -8,7 +8,7 @@ plot_sbc_ecdf <- function(results_coev) {
     theme(strip.background = element_blank())
   # save
   ggsave(
-    filename = here::here("figures/simulation_based_calibration/ecdf.pdf"),
+    filename = "figures/simulation_based_calibration/ecdf.pdf",
     plot = out,
     width = 7,
     height = 5

### 2024-08-27 R/plot_sbc_ecdf_diff.R

@@ -9,7 +9,7 @@ plot_sbc_ecdf_diff <- function(results_coev) {
     theme(strip.background = element_blank())
   # save
   ggsave(
-    filename = here::here("figures/simulation_based_calibration/ecdf_difference.pdf"),
+    filename = "figures/simulation_based_calibration/ecdf_difference.pdf",
     plot = out,
     width = 7,
     height = 5

### 2024-08-27 R/plot_synthetic_example.R

@@ -4,7 +4,7 @@ plot_synthetic_example <- function(fit, delta_theta_promiscuity,
   pA <-
     ggdraw() +
     draw_image(
-      image = here::here("figures/cichlids/cichlids.png"),
+      image = "figures/cichlids/cichlids.png",
       scale = 0.95
       )
   # plot delta theta
@@ -42,7 +42,7 @@ plot_synthetic_example <- function(fit, delta_theta_promiscuity,
     )
   # plot phase plane
   png(
-    here::here("figures/cichlids/phase_plane.png"),
+    "figures/cichlids/phase_plane.png",
     height = 12,
     width = 12,
     units = "cm",
@@ -58,28 +58,28 @@ plot_synthetic_example <- function(fit, delta_theta_promiscuity,
   pC <-
     ggdraw() + 
       draw_image(
-        image = here::here("figures/cichlids/phase_plane.png")
+        image = "figures/cichlids/phase_plane.png"
         ) +
       draw_image(
-        image = here::here("figures/cichlids/low_sperm_size.png"),
+        image = "figures/cichlids/low_sperm_size.png",
         scale = 0.15,
         x = -0.40,
         y = -0.27
         ) +
       draw_image(
-        image = here::here("figures/cichlids/high_sperm_size.png"),
+        image = "figures/cichlids/high_sperm_size.png",
         scale = 0.25,
         x = -0.43,
         y =  0.31
       ) +
       draw_image(
-        image = here::here("figures/cichlids/low_promiscuity.png"),
+        image = "figures/cichlids/low_promiscuity.png",
         scale = 0.15,
         x = -0.25,
         y = -0.41
       ) +
       draw_image(
-        image = here::here("figures/cichlids/high_promiscuity.png"),
+        image = "figures/cichlids/high_promiscuity.png",
         scale = 0.15,
         x =  0.35,
         y = -0.41
@@ -94,7 +94,7 @@ plot_synthetic_example <- function(fit, delta_theta_promiscuity,
     )
   ggsave(
     plot = out,
-    filename = here::here("figures/cichlids/synthetic.pdf"),
+    filename = "figures/cichlids/synthetic.pdf",
     width = 7,
     height = 3
     )

### 2024-08-30 manuscript.Rmd
Update prior for simulations

@@
     number_sections: true
+    includes:
+      in_header: header.tex
 bibliography: references.bib
@@
 linkcolor: blue
-urlcolor: black
+urlcolor: blue
 figsintext : yes
-header-includes:
-  \usepackage{fixltx2e}
-  \usepackage{hyperref}
-  \newcommand\figref[1]{{\hypersetup{linkcolor=blue}\autoref{#1}}}
 ---
@@
 \begin{center}
-Abstract text here...
+Phylogenetic comparative methods are widely used to study the coevolution of
+biological and cultural traits over evolutionary time. Among these methods, 
+some of the most popular are phylogenetic generalised linear mixed models, 
+phylogenetic path analysis, and Pagel's 'discrete' method. While these methods
+have clear benefits under specific scenarios, they also have important
+limitations, such as unrealistic assumptions about the evolutionary process,
+a focus on evolutionary correlation rather than causation, and limits on the 
+number and type of traits that researchers can include. Here, we develop a
+novel class of Bayesian *dynamic phylogenetic models* (DPMs) that overcomes
+these limitations and synthesizes the strengths of existing phylogenetic 
+comparative methods into a single generalizable framework for dynamic
+phylogenetic causal inference. We formally and conceptually introduce this
+class of model, which is applicable to any number of traits following any
+response distribution. We provide a short tutorial for fitting these models
+with the `coevolve` R package, a user-friendly package purpose-built for 
+fitting DPMs in Stan. We highlight previous studies that have used DPMs to 
+make inferences about trait coevolution, including analyses of brain size 
+in non-human primates and social complexity in human societies, and suggest
+some directions for future model development and research.
 \end{center}
@@
 Phylogenetic comparative methods (PCMs) are commonly used to study the
-co-evolution of organismal traits, spanning topics such as anatomy and 
+coevolution of organismal traits, spanning topics such as anatomy and 
 physiology [@Dunn2015; @Garland2005; @Navalon2019; @OConnor2022; @Thayer2018], 
@@
 available [@Garamszegi2014; @Harvey1991; @Nunn2011]. Many PCMs focus on 
-understanding how traits coevolve by estimating evolutionary correlations and 
+understanding *how* traits coevolve by estimating evolutionary correlations and 
 rates of change across phylogenetic trees, which can be used to better 
@@
 evolutionary correlation does not imply evolutionary causation between traits 
-[@Shipley2016]. Thus, PCMs facilitating inference of why traits coevolve and 
+[@Shipley2016]. Thus, PCMs facilitating inference of *why* traits coevolve and 
 change together across time are indispensable tools for testing 
@@
 motivate further methods development. We then formally and conceptually 
-introduce a novel class of dynamic phylogenetic models. We provide a worked 
-synthetic example of coevolution between two continuous traits -- illustrated as
-female promiscuity and sperm length [@Fitzpatrick2009] -- to assess the accuracy
-and uncertainty of inferences made with dynamic phylogenetic models across a 
-reasonable range of sample sizes for phylogenetic research. We also provide an 
-accompanying coding tutorial using our **coevolve** R package to aid empiricists
-in easily applying basic dynamic phylogenetic models to their own datasets. We 
-then demonstrate the generality and flexibility of our method with two empirical
-applications, which extend the model to complex, high-dimensional scenarios 
-using dynamic latent variables. In particular, we show how our method can 
-generate insights into the causal dynamics of trait coevolution across both 
-genetic and cultural evolutionary timescales, using (1) a comparative dataset on
-primate brain size, sociality, diet, and life history traits [@DeCasien2017], 
-which reduced to two coevolving latent variables, and (2) two studies on the 
-evolution of social complexity across a global and a regional sample of 
-pre-industrial human societies [@Ringen_preprint; @Sheehan2023].
+introduce a novel class of *dynamic phylogenetic models* (DPMs). We provide a 
+worked synthetic example of coevolution between two continuous traits -- 
+illustrated as female promiscuity and sperm length [@Fitzpatrick2009] -- to 
+assess the accuracy and uncertainty of inferences made with dynamic phylogenetic
+models across a reasonable range of sample sizes for phylogenetic research. We 
+also provide an accompanying coding tutorial using our `coevolve` R package to
+aid empiricists in easily applying basic dynamic phylogenetic models to their 
+own datasets. We then demonstrate the generality and flexibility of our method 
+with two empirical applications, which extend the model to complex, 
+high-dimensional scenarios using dynamic latent variables. In particular, we 
+show how our method can generate insights into the causal dynamics of trait 
+coevolution across both genetic and cultural evolutionary timescales, using (1) 
+a comparative dataset on primate brain size, sociality, diet, and life history 
+traits [@DeCasien2017], which reduced to two coevolving latent variables, and 
+(2) two studies on the evolution of social complexity across a global and a 
+regional sample of pre-industrial human societies [@Ringen_preprint; 
+@Sheehan2023].
 
@@
 trait correlations among closely related species with similar phenotypes, 
-creating the illusion of convergent co-evolution even when traits evolve 
+creating the illusion of convergent coevolution even when traits evolve 
 independently. Adjustment for phylogeny reduces bias and variance due to 
@@
 over time, with most commonly used methods focusing largely on evolutionary 
-correlation (\figref{fig:dags}a) rather than causation (\figref{fig:dags}b-e).
+correlation (\autoref{fig:dags}a) rather than causation (\autoref{fig:dags}b-e).
 
-```{=latex}
-\begin{figure}
-```
-
-```{r dags, echo=FALSE, out.width='100%'}
-knitr::include_graphics("figures/dags.png")
-```
-
-```{=latex}
-\caption{Explanatory models of trait coevolution. \label{fig:dags}}
-\textit{Note:} Examples of distinct formal approaches to describing and 
-explaining patterns of coevolution among traits (bold letters), with important 
-properties represented in each graph highlighted by blue arrows and text. 
-Approaches range in complexity from (a) simple models of phylogenetic 
-correlations (indicated by bidirectional arrows), which are useful for 
-predictive purposes but fail to explain the causes of trait associations; to 
-(b-e) directed acyclic graph models, which can be used to directly test the 
-causal effects (directed arrows) driving trait associations across a phylogeny. 
-As explained in Box 2 below, explicit causal models are crucial for deciding 
-which traits should be included or excluded from a multivariate analysis to 
-avoid potential biases due to phenomena such as so-called forks (b) and 
-colliders (c). Causal models can also be further distinguished by whether they 
-model relationships among traits as static (b-c) or dynamic (d-e) effects. Only 
-dynamic models can be used to account for feedback processes 
-(t\textsubscript{-1} → t → t\textsubscript{+1}) generated by reciprocal 
-causation (blue arrows) among traits and autoregressive effects within traits 
-(grey arrows) over time (d). For high dimensional problems, it is important to 
-also consider whether inclusion of latent causes into the dynamic DAG (e), 
-capturing dimensions of evolutionary integration among multiple traits, can 
-provide a more parsimonious and theoretically insightful causal model (see 
-\autoref{sec:empirical}).
-\end{figure}
+```{r dags, echo=FALSE, fig.align='center', fig.cap="\\textit{Explanatory models of trait coevolution.} Examples of distinct formal approaches to describing and explaining patterns of coevolution among traits (bold letters), with important properties represented in each graph highlighted by blue arrows and text.
+Approaches range in complexity from (a) simple models of phylogenetic correlations (indicated by bidirectional arrows), which are useful for predictive purposes but fail to explain the causes of trait associations; to (b-e) directed acyclic graph models, which can be used to directly test the causal effects (directed arrows) driving trait associations across a phylogeny.
+Explicit causal models are crucial for deciding which traits should be included or excluded from a multivariate analysis to avoid potential biases due to phenomena such as so-called forks (b) and colliders (c).
+Causal models can also be further distinguished by whether they model relationships among traits as static (b-c) or dynamic (d-e) effects.
+Only dynamic models can be used to account for feedback processes (t\\textsubscript{-1} → t → t\\textsubscript{+1}) generated by reciprocal causation (blue arrows) among traits and autoregressive effects within traits (grey arrows) over time (d).
+For high dimensional problems, it is important to also consider whether inclusion of latent causes into the dynamic DAG (e), capturing dimensions of evolutionary integration among multiple traits, can provide a more parsimonious and theoretically insightful causal model (see \\autoref{sec:empirical}).", out.width='100%'}
+knitr::include_graphics("figures/dags/dags.png")
 ```
@@
 
+Phylogenetic generalized linear (mixed) models are linear regression models used
+to quantify trait covariance between species or populations due to shared 
+evolutionary history (likewise for their predecessor, independent contrasts) 
+[@Blomberg2012; @Grafen1989; @Hadfield2010; @Lynch1991; @Symonds2014]. The most 
+commonly assumed model of trait coevolution in these regression models is 
+Brownian motion (also called a Wiener process, *CITE*), which assumes that the 
+trait variance for a given species or population is proportional to the total 
+branch length of the tree leading from root to tip, and the covariance between 
+any two species is proportional to their amount of shared history. This process
+is often biologically interpreted as reflecting patterns of evolutionary drift 
+(or neutral evolution) across a phylogenetic tree. Empirically, the actual 
+degree of covariance between related species or populations may be less than
+expected under a pure Brownian motion model, such as when evolution is rapid or,
+more mundanely, when measurement error is substantial. In these cases, the 
+"phylogenetic signal" will be weak such that even closely related taxa may have 
+dramatically different trait values [@Blomberg2003; @Kamilar2013], motivating 
+various branch length transformations to the basic model to account for this 
+greater independence among taxa [@Pagel1999]. In cases where traits do not 
+evolve according to a modified Brownian motion process, other adjustments can 
+also be made to the basic phylogenetic generalized linear model to modify the 
+expected rate change in trait (co)variation. For instance, early burst models 
+of adaptive radiation allow the rate of change to decrease over time (*CITE*), 
+while Ornstein-Uhlenbeck models characterize evolutionary change as the product
+of both stochastic and deterministic processes (*CITE*).
+
+Despite their differences, attempts to use parameters from any of these models
+for causal inference share often unstated and undesirable assumptions, namely 
+that there is no selection acting on the traits included as predictors, that all 
+predictor traits are independent and direct causes of the response trait, and 
+that there is no reciprocal causation between traits over time (see 
+\autoref{fig:dags}). These are strong assumptions that are likely to be violated
+in most empirical datasets, limiting the applicability of phylogenetic 
+generalized linear (mixed) models for theory testing and development. In 
+addition, the most commonly used implementation of these models, phylogenetic
+generalized least squares regression, suffers from various other constraints 
+such as being limited to traits with Gaussian errors, overfitting due to a lack
+of parameter regularization, and a failure to accommodate many common data 
+features such as repeated measures of species trait values, missing data, and 
+measurement error. While multilevel/mixed-effects models can be used to address 
+these statistical concerns [@Hadfield2010; @Ives2010; @Ives2011; @Lynch1991; 
+@Martin2020; @Ringen2019], they nonetheless share the same basic limitations 
+with regard to causal inference and are arguably not ideal for modelling 
+adaptation [@Hansen2014] (Box \ref{box:evolutionary-regression-coefficients}).
+
+\mybox[label=box:evolutionary-regression-coefficients]{Evolutionary regression coefficients underestimate causal effects}
+An unintuitive aspect of traditional PCMs is that regression coefficients 
+estimated using cross-sections of contemporary species are used to infer the 
+change in some trait $Y$ in response to another trait $X$ over evolutionary 
+time. The slope of a response trait on a predictor trait is known as the 
+"evolutionary regression coefficient" [@Pagel1993]. For the evolutionary
+regression coefficient to be interpreted as a causal effect, we must, as with 
+all statistical models, assume that there are no unmeasured confounders and that
+our parametric model is correct. But there is an additional caveat: because 
+selection is (generally) a gradual process, the total causal effect of one trait
+on another -- which we define as the change in the optimal trait value of $Y$ as
+a function of the value of $X$ [see @Scholkopf2013 for a closely related 
+definition of causal effects in systems of ordinary differential equations] -- 
+can take a long time to be fully realized. As such, we capture species in the 
+process of adaptation rather than at equilibrium. Consequently, traditional 
+methods underestimate the total causal effect of one trait another. This 
+attenuation bias is a joint product of the strength of selection on the response
+trait and the rate of change in the predictor trait. Intuitively, if one trait 
+changes too quickly, then the other will always be playing catch-up. See 
+\autoref{fig:evo-reg-coef} for an illustration of this process. A solution to 
+this problem is to move beyond traditional regression-based PCMs and explicitly
+model trait change and the adaptive process using dynamic phylogenetic models.
+\endmybox
+
+```{r evo-reg-coef, echo=FALSE, fig.align='center', fig.cap='\\textit{Gradual adaptation flattens the evolutionary regression coefficient.} Ancestral species ("A") evolve towards their optimal trait Y value, which depends on the value of trait X shown on the solid line.
+Ancestors speciate and their descendants ("D") inherit their maladapted trait values, resulting in a slope between the two traits (dashed line) that is flatter than the true relationship.
+Redrawn from Figure 14.4 in Hansen (\\protect\\hyperlink{ref-Hansen2014}{2014}).', out.width='60%'}
+knitr::include_graphics("figures/evolutionary_regression_coefficient.png")
+```
+
 ## Phylogenetic path analysis
+
+Fortunately, a major innovation in statistical research in recent decades has 
+been to realize that even conventional regressions with observational data can 
+be used to infer and test for the presence of causal relationships once 
+appropriate assumptions are encoded into the model structure [@Shipley2016]. 
+This is because specific causal models, usually represented as directed acyclic
+graphs (DAGs, see \autoref{fig:dags}, Box 
+\ref{box:evolutionary-regression-coefficients}), imply specific patterns of 
+correlation (or covariance) among variables [@Pearl2009; @Rubin2005]. An 
+elaborate toolbox now exists to determine the correct test of causal hypotheses
+given a DAG. This has become the state-of-the-art in fields like epidemiology 
+and is increasingly taught and used in evolutionary biology and ecology 
+[@Deffner2022; @McElreath2020; @Shipley2016; @Warrell2020].
+
+If one wants to test for a causal effect of $X$ (say diet) on $Y$ (say brain 
+size), one has to carefully consider the relationship of potential third 
+variables $Z$ (say sociality or life history), which, depending on the structure
+of the DAG, should either be included in the regression analysis or not (see 
+Figure 5). These considerations are what distinguish causal inference from 
+standard predictive applications of multivariate regression. For instance, most 
+researchers know that if $Z$ causes both $X$ and $Y$ (a "fork", 
+\autoref{fig:dags}b), $Z$ should be included to remove spurious associations 
+between $X$ and $Y$. However, if $Z$ is a collider (both $X$ and $Y$ cause $Z$; 
+\autoref{fig:dags}c), $Z$ should not be included in the regression model, as 
+doing so introduces a spurious association between $X$ and $Y$. For example, if
+diet ($X$) and brain size ($Y$) independently caused longevity ($Z$) but were 
+otherwise uncorrelated, conditioning on longevity would introduce a spurious 
+negative association between diet and brain size. This further implies that the
+appropriate set of covariates for testing a causal effect of $X$ on $Y$ may not
+be the same as the set for testing the causal effect of $Z$ on $Y$. 
+Nevertheless, all coefficients in a multivariate regression are commonly
+interpreted as causal effects, a phenomenon known as the "table 2 fallacy" 
+[@Westreich2013]. Most comparative analyses have not yet used the causal 
+inference toolbox and typically include several variables in a phylogenetic 
+generalized least squares model without formal causal justification. This 
+implicitly assumes that all variables are independent causes, which is often 
+problematic.
+
+Phylogenetic path analysis represents a particular application of the causal 
+inference toolbox that overcomes some of the issues outlined above for 
+phylogenetic generalized least squares [@GonzalezVoyer2014; @vonHardenberg2013].
+With phylogenetic path analysis, researchers pre-define specific causal models,
+ideally avoiding colliders and other traps (\autoref{fig:dags}), and compare 
+their fit to the data. Specifically, a DAG implies a set of variables that 
+should be statistically independent, which is then tested using phylogenetic 
+regression, such that several causal hypotheses can be compared. For instance, 
+@Navarrete2016 presented several phylogenetic path analysis models that all 
+included links between life history and social group size with brain size, with
+variable direct or indirect effects of diet and technical intelligence. Thus, 
+phylogenetic path analysis is well-suited for comparing existing hypotheses, 
+with DAGs generated "top-down" from prior theory or empirical literature
+[@GonzalezVoyer2014].
+
+However, the applicability of phylogenetic path analysis is limited to static 
+DAGs lacking any reciprocal effects over time, such as \autoref{fig:dags}b and c 
+where $X$, $Y$, and $Z$ do not have any temporal structure that could be used
+to resolve reciprocal causal effects. Such models assume a constant direction of
+effect between two variables, with bidirectional causal effects only being able
+to be described as unresolved patterns of correlation (\autoref{fig:dags}a). 
+This is an issue because positive (or negative) evolutionary feedback loops are
+predicted by many theoretical models in response to life history tradeoffs and 
+various forms of frequency- and density-dependence, suggesting that reciprocal
+causation is likely to be a common phenomenon explaining patterns of trait 
+coevolution. For example, @McNamara2022 presents a model of the coevolution of
+paternal care and extrapair paternity, which is driven by a frequency-dependent
+feedback loop over evolutionary time between males' tendency to seek extrapair
+copulations and the benefits of their paternal care. The inability to directly
+test for such dynamic feedback effects is a fundamental limitation of 
+phylogenetic path analysis and any other phylogenetic method reliant on static 
+causal models, thus motivating the development of dynamic methods.
 
@@
 
+To date, Pagel's [-@Pagel1994] "discrete" method is the only general application
+of dynamic causal inference to phylogenetic data. By reconstructing the 
+coevolutionary sequence of two traits along the branches of a phylogeny, 
+researchers can infer causality and potentially reciprocal causality through 
+temporal contingency, referred to as "Granger causality" in economics 
+[@Granger1969]. For example, researchers can infer that $X$ evolved first and 
+then made the evolution of $Y$ more likely. This method has resulted in many 
+high-profile publications [@Cornwallis2017; @Fitzpatrick2009; @Kappeler2019; 
+@Sheehan2018; @Shultz2011; @Watts2016], all of which are built on the stronger 
+causal inference of temporal contingency.
+
+Unfortunately, despite its innovative approach to causal inference, Pagel's 
+method is fundamentally limited to investigating the coevolution of binary 
+traits, where dynamic evolutionary transitions can be conceptualized in terms of
+probabilities of switching between two discrete states. Application of this 
+method to evolutionary topics involving continuous phenomena (e.g. morphology,
+life history, brain size, etc.) thus requires that researchers falsely 
+dichotomize naturally occurring variation in traits [e.g. @Cornwallis2017; 
+@Fitzpatrick2009; @Sheehan2018; @Watts2016], which is well-established as an 
+undesirable statistical practice leading to loss of power, biased effect sizes,
+and inconsistent results [@Dawson2012; @Royston2006]. For instance, 
+@Fitzpatrick2009 dichotomized continuously-measured sperm length and speed by 
+classifying traits below the species mean as "low" and those above the species 
+mean as "high", and they collapsed a four-point scale of female promiscuity, 
+based on behavioral and paternity data, into "low" (levels 1 and 2) and "high" 
+(levels 3 and 4). Using Pagel's method, they then showed that sperm evolved to 
+be faster and longer in response to increases in female promiscuity.
+
+The limitation of Pagel's method to binary traits is further exacerbated when 
+studying the (co)evolution of multiple continuous traits, which often motivates
+the incorporation of latent variables into a dynamic causal analysis 
+(\autoref{fig:dags}e). Latent variables are widely used in biology to capture
+theoretically pertinent constructs that are difficult to directly operationalize
+and quantify using a single measurement, such as size and shape dimensions in 
+morphometrics (*CITE*), environmental quality and climate metrics in ecology
+(*CITE*), life history variation [@Stott2024], and canonical axes of 
+correlational selection in quantitative genetics (*CITE*). While latent 
+variables are sometimes used pragmatically as placeholders for deeper causal 
+analysis [@Shipley2016], many evolutionary models directly conceptualize and 
+model causal effects occurring among latent variables. For instance, 
+life-history traits are theorized to evolve along a latent fast-slow continuum 
+[@Bielby2007; @Healy2019; @Stott2024; @Wright2019] that can both cause and in 
+turn be caused by the correlated evolution of a niche complexity continuum 
+reflecting differences in brain size and cognitive ability, social group size, 
+and foraging behavior [@Dunbar2017; @Miller2019; @Navarrete2016] (see 
+\autoref{sec:brain-size-evo}). While prior research has used methods such as 
+phylogenetic principal component analysis (pPCA; *CITE*) to apply Pagel's method
+to latent variables [e.g. @Cornwallis2017], this approach still requires falsely
+dichotomizing latent variables and thus suffers from the same statistical biases
+as with any other continuous measure. Synthesizing continuous and multivariate
+methods with dynamic causal inference thus remains a major challenge for 
+contemporary PCMs.
+
 # Dynamic phylogenetic models
+
+The limitations of current PCMs can be overcome by synthesizing their respective
+strengths into a single generalizable framework for dynamic phylogenetic causal
+inference, applicable to both discrete and continuous traits. We have attempted
+to do so through the development of a novel class of PCMs we refer to simply as
+dynamic phylogenetic models (DPMs) of trait coevolution. Our DPM is driven by a 
+multivariate stochastic differential equation, similar to a multivariate 
+Ornstein-Uhlenbeck (OU) process (see Box \autoref{box:ou-model} for an overview
+of the basic OU model) with greater model flexibility and complexity than is 
+permitted by standard implementations. Our approach arose from two complementary
+goals: first, to extend previous Bayesian implementations of the multivariate OU
+model such that the optimal trait value $\theta$ is updated dynamically (rather
+than assuming piecewise constant, see Box \autoref{box:ou-model}). Second, we 
+wanted a general form for the optimal trait value $\theta$ that could be used to
+assess directionality (X → Y vs. Y → X) and contingencies (X, then Y) in 
+evolution, akin to Pagel's [-@Pagel1994] method but without restrictions on the
+type or number of traits.
+
+\mybox[label=box:ou-model]{The Ornstein-Uhlenbeck (OU) model}
+The OU model is a mean-reverting, stationary Gauss-Markov process. It describes
+change in a trait due to both Gaussian noise and reversion towards some central
+value. In some fields, such as economics, the mean reversion component is called
+"drift" and the stochastic noise is called "diffusion" (i.e., the Vasicek model,
+*CITE*). In evolutionary biology, the mean-reverting quality is loosely 
+interpreted as "selection" and accordingly the Gaussian noise is labeled "drift"
+(*CITE*). The basic OU form is
+
+$$
+dy(t) = \alpha(\theta - y_{t}) + \sigma dW(t)
+$$
+
+where $\alpha$ controls the strength of mean-reverting selection, $\theta$ is 
+the mean trait value, and $\sigma$ controls the strength of drift. When 
+$\alpha = 0$, the model is pure drift Brownian motion. The simplest OU models 
+assume a single evolutionary optimum ($\theta$), or estimate an ancestral 
+optimum along with a global optimum. More elaborate OU models imagine that 
+$\theta$ changes as a function of other variables, turning it into a 
+coevolutionary process with varying selection regimes (i.e., the Hansen model; 
+*CITE*). These approaches exploit the fact that, if selection regimes are 
+piecewise-constant (the optimum is the same within each segment of the 
+phylogenetic tree but allowed to vary at branching points), the OU process can 
+be discretized, giving a closed form solution for the expectation and covariance
+matrix of a trait. Butler and King (*CITE*) provided a maximum-likelihood 
+algorithm for this approach, and a Bayesian implementation was developed by 
+Ross et al. (*CITE*). Depending on the research question, the piecewise constant 
+assumption can be quite restrictive and a poor approximation when tree segments
+are long, due to the assumption of a constant optimum within each discrete tree
+segment.
+
+\endmybox
 
@@
 
+Rather than estimating a phylogenetic variance-covariance matrix as is standard
+for PCMs, we instead partition the deterministic and stochastic components of 
+the coevolutionary process, adapting the continuous-time structural equation 
+modeling approach of @Driver2017 to the phylogenetic context. The evolutionary 
+history of any species or population is modeled as a time series where the 
+deterministic dynamics of an OU process play out over the length of each tree 
+segment, and the stochastic drift components (which are by definition orthogonal
+to selection) are added to the end of each segment as independent samples from 
+the standard normal distribution, scaled by the expected covariance for a given 
+segment duration $t(s)$
+
+\begin{equation}
+  \label{eq:ou-evo-model}
+  d\eta(t) = (\textbf{A}\eta(t) + \textbf{b}) + \textbf{G}dW(t)
+\end{equation}
+
+Here $\eta(t)$ is a vector of the latent variables at time $t$, the matrix 
+$\textbf{A}$ represents "selection" with autoregressive terms on the diagonal
+equivalent to $\alpha$ in the univariate OU process (Box 
+\autoref{box:ou-model}) and off-diagonals representing the effect of each trait
+on the others (e.g., $\textbf{A}[2,1]$ represents the effect of $\eta_1$ on 
+$\eta_2$), and $\textbf{b}$ is a vector of continuous time intercepts that, 
+along with $\textbf{A}$, determine the asymptotic values of $\eta$. The matrix 
+$\textbf{G}$ is the Cholesky decomposition of the "drift" covariance matrix 
+$\textbf{Q}$, such that $\textbf{Q} = \textbf{G}\textbf{G}'$, which scales the
+stochastic Weiner process. The square root of the diagonals in $\textbf{Q}$ are
+equivalent to $\sigma$ in the OU process (Box \autoref{box:ou-model}). Although
+in other types of time-series analyses it is possible to estimate the 
+off-diagonals of $\textbf{Q}$ (i.e., the covariance of the stochastic drift 
+terms), it is not possible to simultaneously estimate them in the phylogenetic
+context while also estimating the off-diagonals of $\textbf{A}$, so in our model
+we assume that they are zero. This is equivalent to the assumption made in 
+Pagel's [-@Pagel1994] method that multiple traits cannot transition together 
+instantaneously.
+
+Following @Driver2017, the solution to \autoref{eq:ou-evo-model} for any time 
+interval $t - t_0$ is
+
+\begin{equation}
+  \label{eq:solution}
+  \begin{aligned}
+    \eta(t) = e^{\textbf{A}(t-t_0)}\eta(t_0)\textbf{A}^{-1}[e^{\textbf{A}
+    (t-t_0)}]\textbf{b} + \int_{t_0}^{t}e^{\textbf{A}(t-s)}\textbf{G}dW(s) \\
+    \text{cov}\biggl[\int_{t_0}^{t}e^{\textbf{A}(t-s)}\textbf{G}dW(s)\biggr] =
+    \text{irow}\biggl(\textbf{A}_\text{\#}^{-1}\Bigl[e^{\textbf{A}_\text{\#}
+    (t-t_0)}- \textbf{I}\Bigr]\text{row}(\textbf{Q})\biggr)
+  \end{aligned}
+\end{equation}
+
+where $\textbf{A}_{\text{\#}}=(\textbf{A}\otimes\textbf{I})+(\textbf{I}\otimes\textbf{A})$, 
+with $\otimes$ denoting the Kronecker product. $\textbf{I}$ is an identity 
+matrix, $\text{row()}$ is an operation that takes elements of a matrix row-wise
+and puts them in a column vector, and $\text{irow()}$ is the inverse of the row
+operation [@Driver2017]. In the OU process, $\frac{dy}{dt}=0$ when $y=\theta$ 
+(Box \autoref{box:ou-model}). Equivalently, we can calculate the equilibrium 
+value $\theta$ for each trait $\eta_i$ in the DPM as
+
+\begin{equation}
+  \label{eq:equilibrium}
+  \theta_{\eta_{i}} = \frac{-(\sum_{j \neq i}\textbf{A}[i,j]\eta_j) + 
+  \textbf{b}_i}{\textbf{A}[i,i]}
+\end{equation}
+
+Note that the vector $\theta_{\eta}$ does not equal the time asymptotic trait 
+values for the system as a whole ($\textbf{b}_{\Delta\infty}$; *CITE*), which 
+are determined by
+
+\begin{equation}
+\label{eq:asymptotic}
+  \textbf{b}_{\Delta\infty} = \textbf{A}^{-1}\textbf{b}
+\end{equation}
+
 ## Implementation
+
+Our strategy for mapping this model onto a phylogenetic tree is described in 
+\autoref{alg:alg} and visualized in \autoref{fig:dpm-algorithm}. In short, we
+iterate over segments of the tree and allow the coevolutionary process to play
+out over the length of each tree segment.
+
+\begin{algorithm}
+  \caption{Implementation of the dynamic phylogenetic model}\label{alg:alg}
+  \DontPrintSemicolon
+  \For{$n \in N$}{
+    Divide the evolutionary history of $n$ into $S$ segments, where each 
+    segment starts with a parent node and ends with a child node or tip.\;
+    Calculate the length of each segment $s \in S$, i.e., duration of time 
+    between the parent and child nodes.\;
+    Initialize the ancestral trait values $\eta_0$.\;
+    \For{$s \in S$}{
+      Solve for $\eta(s)$:
+      \begin{equation}
+        \nonumber
+        \eta(s) = e^{\textbf{A}(t(s))}\eta(s-1)\textbf{A}^{-1}[e^{\textbf{A}
+        (t(s)-t(s-1))}]\textbf{b} + \int_{t(s-1)}^{t(s)}e^{\textbf{A}
+        (t(s-1)-t(s))}\textbf{G}dW(t(s))
+      \end{equation}
+      \;
+    }
+  }
+\end{algorithm}
+
+```{r dpm-algorithm, echo=FALSE, fig.align='center', fig.cap='\\textit{The DPM algorithm.} Example of how our phylogenetic algorithm works (see \\autoref{alg:alg} for formal structure).
+The evolutionary history of each tip (A-E) is described as a time series starting at \\(t(s)\\) and comprised of a set of segments \\(S\\).
+The OU process runs over each segment in order, where the initial trait values for any segment \\(s\\) are the terminal values from \\(s-1\\).
+Each sequence is a combination of shared segments (where tips are evolving together) and unshared segments (where tips are evolving independently).
+For example, S(A) and S(B) start with the same trait values because they both descend directly from the parent segment S(A,B).
+But further evolution along S(A) and S(B) happens independently.
+Dashed lines indicate segment split points.', out.width='100%'}
+knitr::include_graphics("figures/dpm_algorithm.png")
+```
+
+We have implemented this algorithm in a Bayesian inferential framework using the 
+Stan programming language [@Carpenter2017] (see Supplementary Materials for
+example Stan code). Stan employs the computationally efficient Hamiltonian Monte
+Carlo MCMC algorithm to sample from the posterior distribution. Implementing the 
+model in Stan also allows for a greater degree of model flexibility. Users can
+fit the model to any number of traits following any response distribution,
+lifting the limitation of Pagel's [-@Pagel1999] method to binary traits and
+eschewing the need for false dichotimization. Stan also allows for additional 
+features, such as the incorporation of regularization through prior 
+distributions, the inclusion of measurement error on traits, and handling 
+missing data.
+
+## The `coevolve` R package
+
+To aid researchers in applying the DPM to their own data, we have developed the 
+`coevolve` R package (https://github.com/ScottClaessens/coevolve). This 
+fully documented package provides a user-friendly interface for applying DPMs, 
+with simple functions for generating DPM Stan code and data lists, fitting DPMs
+to data using the `cmdstanr` package [@R_cmdstanr], and post-processing and 
+plotting model results.
+
+The development version of the `coevolve` package can be installed and loaded
+using the following R code:
+
+```{r eval=FALSE}
+#install.packages("devtools")
+library(devtools)
+install_github("ScottClaessens/coevolve")
+library(coevolve)
+```
+
+Currently, the `coevolve` package supports binary, ordinal, count, continuous, 
+and positive real traits. The package can also handle missing trait data, 
+repeated observations per taxa, and model comparison using approximate 
+leave-one-out cross-validation. We plan to implement further response 
+distributions and latent variables in future versions of the package, but for 
+now users are encouraged manually adapt the Stan code produced by the package to
+add these features.
 
@@
 
+In order to demonstrate that our formal model is able to accurately recover true
+coevolutionary dynamics, we apply the `coevolve` package to a simple synthetic
+example.
+
+Consider an ancestral species with two continuous traits. As an illustration, 
+let's assume that we are studying the highly diverse cichlid fish of Lake 
+Tanganyika, and the two continuous traits are female promiscuity and sperm size
+(\autoref{fig:synthetic}a). Our hypothesis is that changes in female promiscuity
+will select for changes in sperm size, but not vice versa [@Fitzpatrick2009]. We
+simulate the coevolution of these traits over time, allowing traits to have 
+autoregressive effects (i.e., change in a trait influences itself in the 
+future), coevolutionary effects (i.e., change in a trait influences the other 
+trait in the future), and stochastic drift. In this particular simulation, we 
+specify that changes in female promiscuity influence future changes in sperm 
+size, but not vice versa.
+
+The data generating model is as follows:
+
+\begin{equation}
+  \begin{aligned}
+    \text{spermsize}_t &= (0.90 \times \text{spermsize}_{t-1}) + 
+    (0.85 \times \text{promiscuity}_{t-1}) + \epsilon_{\text{spermsize}} \\
+    \text{promiscuity}_t &= (0.90 \times \text{promiscuity}_{t-1}) + 
+    \epsilon_{\text{promiscuity}} \\
+    \epsilon_{\text{spermsize}}, \epsilon_{\text{promiscuity}} &\sim 
+    \mathcal{N}(0,0.05)
+  \end{aligned}
+\end{equation}
+
+We allow this coevolutionary process to play out over discrete timesteps. In 
+each timestep, there is a small probability ($p = 0.05$) of a speciation event, 
+where the parent species splits into two child species. Coevolutionary dynamics
+then continue independently on each separate phylogenetic branch. We continue 
+this process of evolution and speciation until we have a sufficient sample size
+of cichlid species.
+
+We then use the `coev_fit()` function from the `coevolve` package to fit our 
+statistical model using the resulting phylogenetic tree and the "contemporary" 
+values of female promiscuity and sperm size from the final timestep. The code is
+as follows:
+
+```{r eval=FALSE}
+fit <-
+  coev_fit(
+    data = sim$data,
+    variables = list(
+      Promiscuity = "normal",
+      SpermSize = "normal"
+    ),
+    id = "species",
+    tree = sim$tree
+  )
+```
+
+This function takes a data frame of trait values, a list of variables to 
+coevolve in the model (along with their associated response distributions), the
+column in the dataset that links to the phylogeny, and a phylogeny object of
+class `phylo`. The function generates the code and data list for Stan, compiles
+the model, and fits the model.
+
+We find that, when fitted to data from 100 cichlid species, the statistical 
+model converges normally and is able to accurately recover the true 
+coevolutionary dynamics. The model summary is as follows:
+
+```{r eval=FALSE}
+summary(fit)
+```
+
+```{r fit-summary, echo = FALSE, warning=FALSE}
+summary(tar_read(fit))
+```
+
+```{r delta-theta, echo = FALSE}
+tar_load(delta_theta_promiscuity)
+tar_load(delta_theta_spermsize)
+```
+
+By plugging the fitted model parameters into \autoref{eq:equilibrium} above, 
+we are able to calculate $\Delta\theta_z$, which is defined as the change in the
+equilibrium value of each trait resulting from an absolute deviation increase in
+the other trait. \autoref{fig:synthetic}b visualizes the resulting posterior 
+distributions of $\Delta\theta_z$ for both directions of selection, which we
+can retrieve using the function `coev_calculate_delta_theta()`. For example,
+we might run:
+
+```{r eval=FALSE}
+coev_calculate_delta_theta(
+  object = fit,
+  response = "Promiscuity",
+  predictor = "SpermSize"
+)
+```
+
+We can see from \autoref{fig:synthetic}b that the posterior distribution of 
+$\Delta\theta_z$ for promiscuity → sperm size is positive (median posterior 
+value = `r round(median(delta_theta_spermsize), 2)`, 95% credible interval 
+[`r round(as.numeric(quantile(delta_theta_spermsize, 0.025)), 2)`
+`r round(as.numeric(quantile(delta_theta_spermsize, 0.975)), 2)`])
+suggesting that an absolute deviation increase in female promiscuity results in
+an increase in the equilibrium value of sperm size. However, the posterior 
+distribution of $\Delta\theta_z$ for sperm size → promiscuity includes zero 
+(median posterior value = `r round(median(delta_theta_promiscuity), 2)`, 95% CI
+[`r round(as.numeric(quantile(delta_theta_promiscuity, 0.025)), 2)`
+`r round(as.numeric(quantile(delta_theta_promiscuity, 0.975)), 2)`]) suggesting 
+that an absolute deviation increase in sperm size does not change the 
+equilibrium value of female promiscuity. This result accurately reflects the 
+data generating process.
+
+```{r synthetic, echo=FALSE, fig.align='center', out.width='100%', fig.width=7, fig.height=3, warning=FALSE, fig.cap="\\textit{Synthetic data example.} (a) We use the example of the coevolution of female promiscuity and sperm size in cichlid fish as an illustration.
+Using Pagel's discrete method, Fitzpatrick et al.
+(\\protect\\hyperlink{ref-Fitzpatrick2009}{2009}) dichotomized their initially continuous measures of female promiscuity and sperm size and swimming speed, finding that changes in promiscuity preceded changes in sperm traits.
+(b) Our synthetic data analysis shows that the dynamic phylogenetic model correctly identifies a positive effect of female promiscuity on sperm size, but not of sperm size on promiscuity.
+(c) The flow field shows how continuous trait values of female promiscuity and sperm size are expected to change over evolutionary time as a function of the state of the other variable.
+Black and white fish represent males and females, respectively."}
+tar_read(plot_synthetic)
+```
+
+We can further visualize the influence of female promiscuity on sperm size by 
+plotting a phase plane of the evolutionary dynamics implied by the model. 
+\autoref{fig:synthetic}c depicts the expected change in female promiscuity and 
+sperm size depending on their current states. Both variables have been 
+standardized as z-scores for easier interpretation. The solid lines in the 
+figure represent nullclines where female promiscuity (blue) and sperm size 
+(red) are at equilibrium, depending on the state of the other. This plot is
+easily generated using the following code:
+
+```{r eval=FALSE}
+coev_plot_flowfield(
+  object = fit,
+  var1 = "Promiscuity",
+  var2 = "SpermSize",
+  nullclines = TRUE
+)
+```
+
+The phase plane shows that changes in female promiscuity have large effects on 
+the direction of selection on sperm size. When female promiscuity is low (i.e.,
+mating is predominantly monogamous), investment in large sperm is selected 
+against, but when female promiscuity is high, large sperm are favored. However,
+changes in sperm size do not have the same effect on the direction of selection
+on female promiscuity or the equilibrium value of female promiscuity. That is, 
+changes in sperm size do not affect the evolution of mating systems in our 
+example.
+
+To determine whether this pattern of results generalizes beyond a single 
+simulation run, we iterated the above simulation 100 times for three different 
+sample sizes: 20, 50, and 100 species. Table \@ref{tab:power} shows the 
+proportion of models where the posterior $\Delta\theta_z$ values have 95% 
+credible intervals excluding zero. The results show that the model is relatively 
+underpowered to detect the causal effect of promiscuity on sperm size when there
+are only 20 species, but it is able to detect this causal effect with sufficient
+power when the sample size is 50 or 100.
+
+```{r power, echo=FALSE, warning=FALSE}
+kable(
+  tar_read(power),
+  caption = paste0(
+    "The proportion of models in which the 95% credible interval for ",
+    "$\\Delta\\theta_{z}$ is greater than zero for three different sample ",
+    "sizes. 100 simulations were run for each sample size."
+    ),
+  digits = 2,
+  booktabs = TRUE
+  )
+```
+
+## Simulation-based calibration
+
+In order to further validate our Bayesian model, we ran a simulation-based 
+calibration exercise. Simulation-based calibration is a method to assess whether
+samples of the posterior distribution from a generator (in this case, our Stan
+model) faithfully capture the "true" posterior distribution [@Modrak2023; 
+@Talts2018]. We infer this faithfulness by simulating parameters from the prior
+distribution, using the simulated parameters to simulate multiple datasets, and 
+then fitting the model to those datasets. In a well-calibrated Bayesian model, 
+the resulting samples of the "data-averaged posterior" should be 
+indistinguishable from the prior distribution. If there is a discrepancy between
+the two, this suggests that the model is either misspecified or otherwise 
+inaccurately computing the posterior distribution.
+
+We used the `SBC` R package [@Modrak2023] to run our simulation-based 
+calibration exercise. We generated 500 simulated datasets from the prior ($N$ =
+20, one continuous trait and one binary trait) and fitted the DPM to these
+simulated datasets. We then determined whether the posterior ranks from the 
+fitted models were uniformly distributed by plotting, for several key model 
+parameters and the log-likelihood, the differences between the empirical 
+cumulative distribution function and a perfectly uniform cumulative distribution
+function. If posterior ranks are perfectly uniformly distributed as expected, 
+this should result in a flat line, though some variation from flat is to be 
+anticipated. Ideally, we would expect to see the lines falling within the 
+expected deviations at the 95% level (blue ovals). Indeed, this is what we find
+for our model (\autoref{fig:sbc}), suggesting that the Stan program is 
+well-calibrated.
+
+```{r sbc, echo=FALSE, fig.align='center', out.width='100%', fig.width=7, fig.height=5, warning=FALSE, fig.cap=, fig.cap="\\textit{Simulation-based calibration.} Differences between the empirical cumulative distribution function (ECDF) and a perfectly uniform cumulative distribution function.
+Blue ovals represent theoretical deviations for these values at the 95\\% level."}
+tar_read(plot_sbc2)
+```
+
 ## Empirical applications
-
 \label{sec:empirical}
 
+To illustrate the flexibility of our method, we provide empirical applications 
+that involve several variables ultimately best analyzed using latent variables. 
+In \autoref{sec:brain-size-evo}, we present a novel analysis of existing data on
+primate brain size evolution, while \autoref{sec:social-complexity} summarizes
+two published examples of cultural evolution in humans.
+
 ### Brain size evolution in primates
+\label{sec:brain-size-evo}
+
+Brain size, both in absolute terms and relative to body size, varies 
+tremendously among taxa, with primates as a group having larger brains than 
+other mammals, and humans having the largest brains of all primates [@Boddy2012;
+@Isler2008; @Miller2019; @Smaers2021]. This has motivated a large number of 
+comparative phylogenetic analyses on correlates of brain size, typically framed
+as testing social or ecological benefits or constraints for evolving larger 
+brains [@DeCasien2017; @Dunbar2017; @Isler2009]. However, many have recognized 
+that this approach has reached an impasse, as different hypotheses can be 
+supported depending on the specific dataset [@Powell2017; @Wartel2019] due to 
+variable patterns of missingness and high collinearity among predictors (e.g. 
+diet, social group size, and life history), and reciprocal causation is inherent
+in most theoretical models (e.g. larger brains allow access to better diet which
+removes energetic constraints on larger brains) [@Isler2014]. Add the fact that
+Pagel's discrete method has to our knowledge never been applied to brain size 
+due to its continuous nature, and this field is virtually predisposed to the use
+of DPMs. The issues and innovations highlighted here should also generalize to 
+many other morphological and socio-ecological traits.
+
+To illustrate the potential of DPMs to analyze brain size evolution, we used the
+dataset of @DeCasien2017, augmented by life-history variables from 
+@Herculano2019, and a consensus phylogeny from 10ktrees [@Arnold2010]. We 
+contrasted a typical phylogenetic multiple regression (Fig. 5a), in which all 
+predictors are interpreted as independent causes of brain size, with a latent 
+factor model (Fig. 5b), which was derived in a bottom-up way using exploratory 
+graph analysis (see Supplementary Materials). Note that instead of modeling 
+absolute brain size and adjusting for body size as is typical for phylogenetic
+regression (Fig. 5a), the model in Fig. 5b directly modeled relative brain size
+as the allometric slope for the brain-body relationship; we also accounted for
+variable allometric relationships using random intercepts and slopes at the
+species level, thus allowing species to differ in their brain-body allometry
+[@Smaers2021]. Note also that this DPM used Bayesian imputation rather than 
+removing species with missing data and losing valuable information and 
+statistical power. In contrast to the approach in Fig. 5a, which compares 
+independent causes and might conclude that "primate brain size is predicted by
+diet but not sociality" [@DeCasien2017], the causal structure in Fig 5b groups
+diet, social group size, and relative brain size into a single latent factor, 
+termed here "niche complexity", with another latent factor, "life-history pace",
+with loadings from the variables body size, age at maturity, and longevity. 
+These latent factor groupings are not surprising, given that much work in 
+primate socio-ecology has emphasized connections between diet and group size
+[@Sterck1997; @Wrangham1980], and both are consistently found to correlate with
+relative brain size [@DeCasien2017; @Dunbar2017]; the integration of life 
+history traits and body size is also well established (*CITE*).
+
+A DPM analyzing the coevolution of these two latent factors (see Supplementary
+Materials) showed that evolutionary changes in life-history pace (LH) cause 
+evolutionary changes in niche complexity (NC), but not vice versa (Fig. 5c). In
+other words, evolution of larger size and slower life history was inferred to 
+precede increases in relative brain size, group size, and frugivory. This is 
+consistent with the observations that repeated changes in size characterized
+mammalian evolution as lineages radiated into new niches [@Pagel2022] and 
+changes in body size associated with new niches were arguably primarily 
+responsible for changes in relative brain size [@Smaers2021]. Further, debates
+about the relative importance of diet and sociality in brain size evolution 
+[@DeCasien2017; @Dunbar2017] seem to be moot as both are sides of the same 
+socio-ecological coin intimately tied to relative brain size [see also 
+@Powell2017; @Wartel2019]. Many extensions of this model are possible, and more
+theoretically motivated causal models should be explored to further advance our
+understanding of brain size evolution.
 
 ### Evolution of social complexity across human societies
+\label{sec:social-complexity}
+
+Human societies are currently and have historically been tremendously diverse, 
+ranging from nomadic hunter-gatherers to large-scale states or empires. Many 
+anthropologists have used comparative approaches to better understand this 
+diversity [@Hooper2024; @Nunn2011], including many applications of Pagel's 
+discrete method [@Sheehan2018; @Watts2016]. Here we summarize two studies that
+have used our DPM.
+
+@Ringen_preprint examined the rise of social complexity using a global sample of
+186 pre-industrial societies that ranged from hunter-gatherers to agrarian 
+empires (Figure 6A). They included nine measures of complexity and three 
+measures of subsistence (Figure 6B), and compared three competing static causal
+models for the relationships between these variables: an "agricentric" model 
+wherein agricultural intensification causes all other variables, a single factor
+model in which a latent "complexity" factor explains all observed variables 
+[as in @Turchin2018], and a two-factor model [as proposed by @Chick1997]. Model
+comparison supported the two-factor model (the correlation structure of which is
+shown in Figure 5B). The authors labeled one factor "resource-use 
+intensification" (RI) and the other "technological and social differentiation" 
+(TSD). A DPM analyzing the co-evolution of the two latent variables RI and TSD 
+showed that, while these two variables are highly correlated (Figure 6C left),
+evolutionary changes in RI were inferred to cause changes in TSD but not vice
+versa (Figure 6C right). In other words, subsistence intensification was a 
+leader, not a follower, in the evolution of complex societies. These results 
+contrast with and complement @Turchin2018, who found a single dimension of 
+complexity (though in a mostly agricultural sample) and @Sheehan2018, who found
+reciprocal causation between agricultural intensification and political 
+complexity (though in a more regional sample).
+
+@Sheehan2023 studied the coevolution of religious and political authority in 97
+Austronesian societies. Religious and political authority were both coded as 
+four-level ordinal variables: absent, sub-local authority, local authority, and
+supra-local authority. Mapping these variables onto a phylogeny of Austronesian
+languages revealed that both religious and political authority had high 
+phylogenetic signals, suggesting that a DPM could reasonably be used to assess
+the coevolution of these variables over cultural evolutionary time. Instead of
+dichotomizing the two ordinal variables to use Pagel's discrete method, as 
+previous work had done [@Sheehan2018; @Watts2016], the authors explicitly 
+modelled both variables as ordinal in a DPM. The model revealed that both 
+religious and political authority coevolved reciprocally over time. In other 
+words, increases in religious authority led to strong positive selection on 
+political authority and, likewise, increases in political authority led to 
+strong positive selection on religious authority. This is akin to "runaway
+selection" where both traits enter a positive feedback loop and increase 
+concomitantly. This coevolutionary relationship makes sense in light of 
+Austronesian ethnographies, which describe how both forms of authority are 
+tightly interdependent and have often historically served to legitimize one 
+another [e.g. @Goodenough2002].
 
@@
 
+xxx
+
 # References

### 2024-08-30 R/compile_model.R

@@ -17,7 +17,8 @@ compile_model <- function() {
         SpermSize   = "normal"
       ),
       id = "id",
-      tree = tree
+      tree = tree,
+      prior = list(A_offdiag = "normal(0, 2)")
     )
   # write stan code to file
   file <- 

### 2024-08-30 R/packages.R

@@ -12,8 +12,8 @@ tar_option_set(
   #  name = "dpm",
   #  workers = 50,
   #  script_lines = "module load R",
-  #  slurm_time_minutes = 20,
-  #  slurm_memory_gigabytes_per_cpu = 2,
+  #  slurm_time_minutes = 60*12,
+  #  slurm_memory_gigabytes_per_cpu = 8,
   #  slurm_cpus_per_task = 4
   #)
 )

### 2024-08-30 manuscript.Rmd
Update manuscript

@@
 some of the most popular are phylogenetic generalised linear mixed models, 
-phylogenetic path analysis, and Pagel's 'discrete' method. While these methods
+phylogenetic path analysis, and Pagel's `discrete' method. While these methods
 have clear benefits under specific scenarios, they also have important
@@
 number and type of traits that researchers can include. Here, we develop a
-novel class of Bayesian *dynamic phylogenetic models* (DPMs) that overcomes
-these limitations and synthesizes the strengths of existing phylogenetic 
-comparative methods into a single generalizable framework for dynamic
-phylogenetic causal inference. We formally and conceptually introduce this
-class of model, which is applicable to any number of traits following any
+novel class of Bayesian \textit{dynamic phylogenetic models} (DPMs) that
+overcomes these limitations and synthesizes the strengths of existing 
+phylogenetic comparative methods into a single generalizable framework for 
+dynamic phylogenetic causal inference. We formally and conceptually introduce
+this class of model, which is applicable to any number of traits following any
 response distribution. We provide a short tutorial for fitting these models
-with the `coevolve` R package, a user-friendly package purpose-built for 
+with the \texttt{coevolve} R package, a user-friendly package purpose-built for 
 fitting DPMs in Stan. We highlight previous studies that have used DPMs to 
@@
 \begin{equation}
+  \label{eq:generative}
   \begin{aligned}
@@
 underpowered to detect the causal effect of promiscuity on sperm size when there
-are only 20 species, but it is able to detect this causal effect with sufficient
-power when the sample size is 50 or 100.
+are only 20 or 50 species, but it is able to detect this causal effect with 
+sufficient power when the sample size is 100. Note, however, that these power
+estimates apply only to the generative model in \autoref{eq:generative};
+we would expect to find different power estimates when simulating with different
+effect sizes and variables with different response distributions.
 

### 2024-08-30 R/plot_synthetic_example.R

@@ -24,9 +24,10 @@ plot_synthetic_example <- function(fit, delta_theta_promiscuity,
     stat_slab(alpha = 0.8) +
     scale_x_continuous(
       name = expression(paste(Delta, theta[z])),
-      limits = c(-5, 15),
+      limits = c(-15, 15),
       expand = c(0, 0)
     ) +
+    ylim(c(0, 1.2)) +
     scale_fill_manual(values = c("#c55852", "#5387b6")) +
     theme_classic() +
     theme(

### 2024-08-30 R/plot_synthetic_example.R
Update synthetic plot

@@ -24,10 +24,10 @@ plot_synthetic_example <- function(fit, delta_theta_promiscuity,
     stat_slab(alpha = 0.8) +
     scale_x_continuous(
       name = expression(paste(Delta, theta[z])),
-      limits = c(-15, 15),
+      limits = c(-10, 15),
       expand = c(0, 0)
     ) +
-    ylim(c(0, 1.2)) +
+    ylim(c(0, 1.3)) +
     scale_fill_manual(values = c("#c55852", "#5387b6")) +
     theme_classic() +
     theme(

### 2024-09-05 manuscript.Rmd
Update manuscript.Rmd

@@
 magic fix for all sources of unobserved confounding, nor does it guarantee that 
-resulting estimates are causally interpretable (*CITE*). Therefore, while all 
+resulting estimates are causally interpretable [@Hansen2014]. Therefore, while all 
 PCMs employ statistical techniques that reduce bias, these methods vary widely 

### 2024-09-05 manuscript.Rmd
Update manuscript.Rmd
Added draft of conclusions and limitations

@@
 
-# Conclusion
+# Conclusions and Limitations
 
-xxx
+First paragraph repeats major advances of the method:
+Dynamic model, analogous to Pagel’s discrete, but without being limited to binary traits
+Can handle any distribution, latent factors
+Can detect evolutionary contingencies (A→B, B→A) and reciprocal causation
+Unbiased estimator, empirical examples summary
+
+Second paragraph states limitations, starting with specific to this method then going broader
+1: Correlated drift cannot be separated from cross-selection (though this is possible with ‘real’ longitudinal data) - biological examples of traits under correlated drift?
+2: Cross-selection effects assumed to be linear - this can be somewhat relaxed with clade-specific parameters, but still locally linear
+3: Computational cost - as with any Bayesian model - can range from minutes (for small phylogenies / datasets) to hours (SCCS example) or even days (for the most complex models, or if looped over a sample of trees)
+4: Interpretation - we’re detecting contingency, not true causality; though this is the same for other kinds of causal inference (Pearl, Pagel, Turchin)
+5: Data quality and inference - comparative data tend to be crude, and all comparative analyses assume that extant variation reflects all past variation.
+Virtual archeology / paleontology is not as good as real archeology / paleontology!
+This can be somewhat amended by including information about ancestral nodes, e.g.
+complexity SCCS analysis put prior for low complexity on last common ancestor - in principle this could be done for any internal node, with prior reflecting exact state of knowledge
 

### 2024-09-05 manuscript.Rmd
Update manuscript.Rmd

@@
 ---
-title: "Causal inference and the detection of coevolutionary contingencies using dynamic phylogenetic models"
+title: "Causal inference and the detection of coevolutionary contingencies using generalized dynamic phylogenetic models"
 author: "E. Ringen, J. S. Martin, S. Claessens, & A. V. Jaeggi"
@@
 number and type of traits that researchers can include. Here, we develop a
-novel class of Bayesian \textit{dynamic phylogenetic models} (DPMs) that
+novel class of Bayesian \textit{generalized dynamic phylogenetic models} (GDPMs) that
 overcomes these limitations and synthesizes the strengths of existing 
@@
 with the \texttt{coevolve} R package, a user-friendly package purpose-built for 
-fitting DPMs in Stan. We highlight previous studies that have used DPMs to 
+fitting GDPMs in Stan. We highlight previous studies that have used GDPMs to 
 make inferences about trait coevolution, including analyses of brain size 
@@
 motivate further methods development. We then formally and conceptually 
-introduce a novel class of *dynamic phylogenetic models* (DPMs). We provide a 
+introduce a novel class of *generalized dynamic phylogenetic models* (GDPMs). We provide a 
 worked synthetic example of coevolution between two continuous traits -- 
@@
 type II (false-negative), type M (magnitude), and type S (sign) errors during 
-statistical inference (*CITE*). Nonetheless, adjustment for phylogeny is not a 
+statistical inference [@felsenstein1985phylogenies; @harvey1998phylogenetic]. Nonetheless, adjustment for phylogeny is not a 
 magic fix for all sources of unobserved confounding, nor does it guarantee that 
@@
 commonly assumed model of trait coevolution in these regression models is 
-Brownian motion (also called a Wiener process, *CITE*), which assumes that the 
-trait variance for a given species or population is proportional to the total 
+Brownian motion, which assumes that the trait variance for a given species or population is proportional to the total 
 branch length of the tree leading from root to tip, and the covariance between 
-any two species is proportional to their amount of shared history. This process
+any two species is proportional to their amount of shared history [@felsenstein1985phylogenies]. This process
 is often biologically interpreted as reflecting patterns of evolutionary drift 
@@
 expected rate change in trait (co)variation. For instance, early burst models 
-of adaptive radiation allow the rate of change to decrease over time (*CITE*), 
+of adaptive radiation allow the rate of change to decrease over time [@harmon2010early], 
 while Ornstein-Uhlenbeck models characterize evolutionary change as the product
-of both stochastic and deterministic processes (*CITE*).
+of both stochastic and deterministic processes [@Hansen2014; @butler2004phylogenetic].
 
@@
 
-To date, Pagel's [-@Pagel1994] "discrete" method is the only general application
+To date, Pagel's [-@Pagel1994] "discrete" method is the most widely-used application
 of dynamic causal inference to phylogenetic data. By reconstructing the 
@@
 and quantify using a single measurement, such as size and shape dimensions in 
-morphometrics (*CITE*), environmental quality and climate metrics in ecology
-(*CITE*), life history variation [@Stott2024], and canonical axes of 
-correlational selection in quantitative genetics (*CITE*). While latent 
+morphometrics [@zelditch2012geometric], environmental quality and climate metrics in ecology
+[@arhonditsis2006exploring], life history variation [@Stott2024], and canonical axes of 
+correlational selection in quantitative genetics [@posthuma2003theory]. While latent 
 variables are sometimes used pragmatically as placeholders for deeper causal 
@@
 \autoref{sec:brain-size-evo}). While prior research has used methods such as 
-phylogenetic principal component analysis (pPCA; *CITE*) to apply Pagel's method
+phylogenetic principal component analysis to apply Pagel's method
 to latent variables [e.g. @Cornwallis2017], this approach still requires falsely
@@
 
-# Dynamic phylogenetic models
+## Dynamic methods with continuous traits
 
-The limitations of current PCMs can be overcome by synthesizing their respective
-strengths into a single generalizable framework for dynamic phylogenetic causal
-inference, applicable to both discrete and continuous traits. We have attempted
-to do so through the development of a novel class of PCMs we refer to simply as
-dynamic phylogenetic models (DPMs) of trait coevolution. Our DPM is driven by a 
-multivariate stochastic differential equation, similar to a multivariate 
-Ornstein-Uhlenbeck (OU) process (see Box \autoref{box:ou-model} for an overview
-of the basic OU model) with greater model flexibility and complexity than is 
-permitted by standard implementations. Our approach arose from two complementary
-goals: first, to extend previous Bayesian implementations of the multivariate OU
-model such that the optimal trait value $\theta$ is updated dynamically (rather
-than assuming piecewise constant, see Box \autoref{box:ou-model}). Second, we 
-wanted a general form for the optimal trait value $\theta$ that could be used to
-assess directionality (X → Y vs. Y → X) and contingencies (X, then Y) in 
-evolution, akin to Pagel's [-@Pagel1994] method but without restrictions on the
-type or number of traits.
-
-\mybox[label=box:ou-model]{The Ornstein-Uhlenbeck (OU) model}
+There are several other approaches to dynamic causal inference, based on the Ohrnstein-Uhlenbeck (OU) model, that relax Pagel's assumption of discrete traits. 
 The OU model is a mean-reverting, stationary Gauss-Markov process. It describes
@@
 "drift" and the stochastic noise is called "diffusion" (i.e., the Vasicek model,
-*CITE*). In evolutionary biology, the mean-reverting quality is loosely 
+[@vasicek1977equilibrium]). In evolutionary biology, the mean-reverting quality is loosely 
 interpreted as "selection" and accordingly the Gaussian noise is labeled "drift"
-(*CITE*). The basic OU form is
+[@uhlenbeck1930theory; @hansen1997stabilizing; @Hansen2014]. The basic OU form is
 
@@
 coevolutionary process with varying selection regimes (i.e., the Hansen model; 
-*CITE*). These approaches exploit the fact that, if selection regimes are 
+[@hansen1997stabilizing]). These approaches exploit the fact that, if selection regimes are 
 piecewise-constant (the optimum is the same within each segment of the 
@@
 be discretized, giving a closed form solution for the expectation and covariance
-matrix of a trait. Butler and King (*CITE*) provided a maximum-likelihood 
+matrix of a trait. Butler and King [@butler2004phylogenetic] provided a maximum-likelihood 
 algorithm for this approach, and a Bayesian implementation was developed by 
-Ross et al. (*CITE*). Depending on the research question, the piecewise constant 
-assumption can be quite restrictive and a poor approximation when tree segments
-are long, due to the assumption of a constant optimum within each discrete tree
-segment.
+Ross et al. [@ross2016origins]. These approaches assume that one trait influences the evolution of another, 
+but not vice versa (i.e.
+no feedback).
+The multivariate OU model was further extended to allow the estimation of reciprocal, directed evolution between traits [@bartoszek2012phylogenetic], a method that has been implemented in the mvslouch package [@bartoszek2024fast].
+This approach aligns closest with our goals, however the current maximum-likelihood implementation only allows for Gaussian traits, suffers from estimation problems, and lacks flexibility (such as accounting for latent factors, measurement error, missing data, repeated measures, etc.), which is why what motivated us to pursue a Bayesian implementation in the Stan probabilistic programming language [@Carpenter2017].
 
-\endmybox
+# Dynamic phylogenetic models
+
+The limitations of current PCMs can be overcome by synthesizing their respective
+strengths into a single generalizable framework for dynamic phylogenetic causal
+inference, applicable to both discrete and continuous traits. We have attempted
+to do so through the development of what we refer to simply as generalized
+dynamic phylogenetic models (GDPMs) of trait coevolution. Our GDPM is driven by a 
+multivariate stochastic differential equation, similar to a multivariate 
+Ornstein-Uhlenbeck (OU) process (see section 2.4. for an overview
+of the basic OU model) with greater model flexibility and complexity than is 
+permitted by current implementations. Our approach arose from two complementary
+goals: first, to extend previous Bayesian implementations of the multivariate OU
+model such that the optimal trait value $\theta$ is updated dynamically (rather
+than assuming piecewise constant, see 2.4.). Second, we 
+wanted a general form for the optimal trait value $\theta$ that could be used to
+assess directionality (X → Y vs. Y → X) and contingencies (X, then Y) in 
+evolution, akin to Pagel's [-@Pagel1994] method and existing OU methods [@bartoszek2012phylogenetic; @bartoszek2024fast] but without restrictions on the
+type or number of traits, and with more flexibility to account for the realities of comparative data.
 
@@
 (Box \autoref{box:ou-model}). Equivalently, we can calculate the equilibrium 
-value $\theta$ for each trait $\eta_i$ in the DPM as
+value $\theta$ for each trait $\eta_i$ in the GDPM as
 
@@
 Note that the vector $\theta_{\eta}$ does not equal the time asymptotic trait 
-values for the system as a whole ($\textbf{b}_{\Delta\infty}$; *CITE*), which 
+values for the system as a whole ($\textbf{b}_{\Delta\infty}$), which 
 are determined by
@@
 \begin{algorithm}
-  \caption{Implementation of the dynamic phylogenetic model}\label{alg:alg}
+  \caption{Implementation of the generalized dynamic phylogenetic model}\label{alg:alg}
   \DontPrintSemicolon
@@
 
-```{r dpm-algorithm, echo=FALSE, fig.align='center', fig.cap='\\textit{The DPM algorithm.} Example of how our phylogenetic algorithm works (see \\autoref{alg:alg} for formal structure).
+```{r GDPM-algorithm, echo=FALSE, fig.align='center', fig.cap='\\textit{The GDPM algorithm.} Example of how our phylogenetic algorithm works (see \\autoref{alg:alg} for formal structure).
 The evolutionary history of each tip (A-E) is described as a time series starting at \\(t(s)\\) and comprised of a set of segments \\(S\\).
@@
 
-To aid researchers in applying the DPM to their own data, we have developed the 
+To aid researchers in applying the GDPM to their own data, we have developed the 
 `coevolve` R package (https://github.com/ScottClaessens/coevolve). This 
-fully documented package provides a user-friendly interface for applying DPMs, 
-with simple functions for generating DPM Stan code and data lists, fitting DPMs
+fully documented package provides a user-friendly interface for applying GDPMs, 
+with simple functions for generating GDPM Stan code and data lists, fitting GDPMs
 to data using the `cmdstanr` package [@R_cmdstanr], and post-processing and 
@@
 calibration exercise. We generated 500 simulated datasets from the prior ($N$ =
-20, one continuous trait and one binary trait) and fitted the DPM to these
+20, one continuous trait and one binary trait) and fitted the GDPM to these
 simulated datasets. We then determined whether the posterior ranks from the 
@@
 due to its continuous nature, and this field is virtually predisposed to the use
-of DPMs. The issues and innovations highlighted here should also generalize to 
+of GDPMs. The issues and innovations highlighted here should also generalize to 
 many other morphological and socio-ecological traits.
 
-To illustrate the potential of DPMs to analyze brain size evolution, we used the
+To illustrate the potential of GDPMs to analyze brain size evolution, we used the
 dataset of @DeCasien2017, augmented by life-history variables from 
@@
 species level, thus allowing species to differ in their brain-body allometry
-[@Smaers2021]. Note also that this DPM used Bayesian imputation rather than 
+[@Smaers2021]. Note also that this GDPM used Bayesian imputation rather than 
 removing species with missing data and losing valuable information and 
@@
 relative brain size [@DeCasien2017; @Dunbar2017]; the integration of life 
-history traits and body size is also well established (*CITE*).
+history traits and body size is also well established [@Healy2019; @Wright2019].
 
-A DPM analyzing the coevolution of these two latent factors (see Supplementary
+A GDPM analyzing the coevolution of these two latent factors (see Supplementary
 Materials) showed that evolutionary changes in life-history pace (LH) cause 
@@
 discrete method [@Sheehan2018; @Watts2016]. Here we summarize two studies that
-have used our DPM.
+have used our GDPM.
 
@@
 intensification" (RI) and the other "technological and social differentiation" 
-(TSD). A DPM analyzing the co-evolution of the two latent variables RI and TSD 
+(TSD). A GDPM analyzing the co-evolution of the two latent variables RI and TSD 
 showed that, while these two variables are highly correlated (Figure 6C left),
@@
 languages revealed that both religious and political authority had high 
-phylogenetic signals, suggesting that a DPM could reasonably be used to assess
+phylogenetic signals, suggesting that a GDPM could reasonably be used to assess
 the coevolution of these variables over cultural evolutionary time. Instead of
@@
 previous work had done [@Sheehan2018; @Watts2016], the authors explicitly 
-modelled both variables as ordinal in a DPM. The model revealed that both 
+modelled both variables as ordinal in a GDPM. The model revealed that both 
 religious and political authority coevolved reciprocally over time. In other 
@@
 
+We presented 
 First paragraph repeats major advances of the method:

### 2024-09-05 manuscript.Rmd
updated abstract

@@
 \begin{center}
-Phylogenetic comparative methods are widely used to study the coevolution of
-biological and cultural traits over evolutionary time. Among these methods, 
-some of the most popular are phylogenetic generalised linear mixed models, 
-phylogenetic path analysis, and Pagel's `discrete' method. While these methods
-have clear benefits under specific scenarios, they also have important
-limitations, such as unrealistic assumptions about the evolutionary process,
-a focus on evolutionary correlation rather than causation, and limits on the 
-number and type of traits that researchers can include. Here, we develop a
-novel class of Bayesian \textit{generalized dynamic phylogenetic models} (GDPMs) that
-overcomes these limitations and synthesizes the strengths of existing 
-phylogenetic comparative methods into a single generalizable framework for 
-dynamic phylogenetic causal inference. We formally and conceptually introduce
-this class of model, which is applicable to any number of traits following any
-response distribution. We provide a short tutorial for fitting these models
-with the \texttt{coevolve} R package, a user-friendly package purpose-built for 
-fitting GDPMs in Stan. We highlight previous studies that have used GDPMs to 
-make inferences about trait coevolution, including analyses of brain size 
-in non-human primates and social complexity in human societies, and suggest
-some directions for future model development and research.
+Phylogenetic comparative methods are widely used to study the coevolution of biological or cultural traits over evolutionary time.
+Among the most popular of these methods are phylogenetic generalised linear mixed models, phylogenetic path analysis, Ornstein-Uhlenbeck models, and Pagel's `discrete' method.
+While these methods are appropriate under specific scenarios, they also limited by often unrealistic assumptions about the evolutionary process and limits on the  number and type of traits that researchers can include.
+Here, we develop a novel class of \textit{generalized dynamic phylogenetic models} (GDPMs) that overcomes these limitations and synthesizes the strengths of existing methods into a generalizable framework for dynamic phylogenetic inference.
+Treating the phylogeny as an implicit time series, GDPMs model the co-evolution of any number of traits undergoing both selection and drift, capable of inferring directed evolution (X -> Y vs Y -> X), feedback (X <-> Y), and contigencies (e.g., first X, then Y).
+We introduce the \texttt{coevolve} R package, a user-friendly package purpose-built for fitting GDPMs in a Bayesian framework using Stan.
+We provide a short tutorial for fitting these models and present several empirical applications of GDPMs to make inferences about trait coevolution, including analyses of brain size in non-human primates and social complexity in human societies.
+ADD BRIEF SUMMARY/HIGHLIGHTS OF THE EMPIRICAL RESULTS.
 \end{center}

### 2024-09-05 manuscript.Rmd
Update manuscript.Rmd
finalized abstract, spelled out conclusions and limitations

@@
 Treating the phylogeny as an implicit time series, GDPMs model the co-evolution of any number of traits undergoing both selection and drift, capable of inferring directed evolution (X -> Y vs Y -> X), feedback (X <-> Y), and contigencies (e.g., first X, then Y).
-We introduce the \texttt{coevolve} R package, a user-friendly package purpose-built for fitting GDPMs in a Bayesian framework using Stan.
-We provide a short tutorial for fitting these models and present several empirical applications of GDPMs to make inferences about trait coevolution, including analyses of brain size in non-human primates and social complexity in human societies.
-ADD BRIEF SUMMARY/HIGHLIGHTS OF THE EMPIRICAL RESULTS.
+We introduce the \texttt{coevolve} R package, a user-friendly package purpose-built for fitting GDPMs in a Bayesian framework using Stan, accompanied by a short tutorial.
+We use synthetic data and simulation-based calibration to show that our models can recover true parameter values and yield unbiased estimates.
+We also present several empirical applications of GDPMs, including analyses of brain size, life history and socio-ecology in non-human primates and social complexity in human societies.
+These examples highlight the flexibility of the Bayesian implementation, allowing researchers to model latent variables, mulitlevel structures and repeated measures, measurement error, missing data, and other complexities inherent in comparative data.
 \end{center}
@@
 
-# Conclusions and Limitations
+# Conclusions and Limitations (ROUGH DRAFT!)
 
-We presented 
-First paragraph repeats major advances of the method:
-Dynamic model, analogous to Pagel’s discrete, but without being limited to binary traits
-Can handle any distribution, latent factors
-Can detect evolutionary contingencies (A→B, B→A) and reciprocal causation
-Unbiased estimator, empirical examples summary
+We presented a novel class of phylogenetic methods that allows the detection of coevolutionary dynamics and has several advantages over existing approaches.
+These include an explicitly dynamic approach, which can detect evolutionary contingencies and reciprocal causation, coupled with the enormous flexibility of the Bayesian implementation in the Stan programming language, allowing researchers to naturally account for latent factors, repeated measures, measurement error, missing data, and other real-life complexities of comparative data.
+The coevolve R package supports some of these complexities (such as missing data), as well as a reasonable diversity of distributions, facilitating the adoption of this method for most R users, while some researchers might prefer the flexibility of coding their models directly in Stan.
 
-Second paragraph states limitations, starting with specific to this method then going broader
-1: Correlated drift cannot be separated from cross-selection (though this is possible with ‘real’ longitudinal data) - biological examples of traits under correlated drift?
-2: Cross-selection effects assumed to be linear - this can be somewhat relaxed with clade-specific parameters, but still locally linear
-3: Computational cost - as with any Bayesian model - can range from minutes (for small phylogenies / datasets) to hours (SCCS example) or even days (for the most complex models, or if looped over a sample of trees)
-4: Interpretation - we’re detecting contingency, not true causality; though this is the same for other kinds of causal inference (Pearl, Pagel, Turchin)
-5: Data quality and inference - comparative data tend to be crude, and all comparative analyses assume that extant variation reflects all past variation.
-Virtual archeology / paleontology is not as good as real archeology / paleontology!
-This can be somewhat amended by including information about ancestral nodes, e.g.
-complexity SCCS analysis put prior for low complexity on last common ancestor - in principle this could be done for any internal node, with prior reflecting exact state of knowledge
+Despite several advances, our method is still subject to several limitations, some specific to GDPMs, and some more broadly applicable.
+In cases where traits experience correlated drift, GDPMs will struggle to separate this effect from cross-selection (EXPAND ON THIS, PERHAPS WITH A REAL BIOLOGICAL EXAMPLE?).
+GDPMs assume that cross-selection effects are linear; this assumption can be somewhat relaxed by using clade-specific parameters, though the effects will still be 'locally' linear (i.e.
+within a given clade).
+As with any Bayesian estimation, GDPMs can be computationally costly, though we note that the Hamiltonian Monte Carlo sampler of Stan is substantially more efficient than conventional MCMC samplers, typically requiring only a few thousand instead of millions of iterations; the models presented in this paper generally took minutes to hours to run.
+As with all comparative analyses, GDPMs are subject to concerns about data quality and bias; e.g.
+that extant trait variation reflects past variation, that available phylogenies accurately reflect true evolutionary history, etc.
+We also note that since GDPMs work as a time series, inference or knowledge about ancestral states is particularly crucial, and relevant information should be incorporated into the models - for instance, Ringen et al [@Ringen_preprint] used priors for low complexity on the last common ancestor of all human societies; in this way, virtual archeology (or paleontology) can be enhanced by real archeology.
+Lastly, we caution that GDPMs detect contingencies, and not causation per se, though this caveat applies to all approaches to causal inference from observational data [@Pearl2009; @Pagel1994].
 

### 2024-09-05 manuscript.Rmd
Update manuscript.Rmd

@@
 title: "Causal inference and the detection of coevolutionary contingencies using generalized dynamic phylogenetic models"
-author: "E. Ringen, J. S. Martin, S. Claessens, & A. V. Jaeggi"
+author: "E. Ringen, S. Claessens, J. S. Martin, & A. V. Jaeggi"
 date: "`r Sys.Date()`"
@@
 Phylogenetic comparative methods are widely used to study the coevolution of biological or cultural traits over evolutionary time.
-Among the most popular of these methods are phylogenetic generalised linear mixed models, phylogenetic path analysis, Ornstein-Uhlenbeck models, and Pagel's `discrete' method.
-While these methods are appropriate under specific scenarios, they also limited by often unrealistic assumptions about the evolutionary process and limits on the  number and type of traits that researchers can include.
+Among the most popular of these methods are phylogenetic generalised linear mixed models, phylogenetic path analysis, and Pagel's `discrete' method.
+While these methods are appropriate under specific scenarios, they are also limited by often unrealistic assumptions about the evolutionary process and limits on the number and type of traits that researchers can include.
 Here, we develop a novel class of \textit{generalized dynamic phylogenetic models} (GDPMs) that overcomes these limitations and synthesizes the strengths of existing methods into a generalizable framework for dynamic phylogenetic inference.
@@
 We use synthetic data and simulation-based calibration to show that our models can recover true parameter values and yield unbiased estimates.
-We also present several empirical applications of GDPMs, including analyses of brain size, life history and socio-ecology in non-human primates and social complexity in human societies.
-These examples highlight the flexibility of the Bayesian implementation, allowing researchers to model latent variables, mulitlevel structures and repeated measures, measurement error, missing data, and other complexities inherent in comparative data.
+We also present several empirical applications of GDPMs, including analyses of brain size in non-human primates and social complexity in human societies.
+These examples highlight the flexibility of the Bayesian framework, allowing researchers to model latent variables, mulitlevel structures and repeated measures, measurement error, missing data, and other complexities inherent in comparative data.
 \end{center}
@@
 We presented a novel class of phylogenetic methods that allows the detection of coevolutionary dynamics and has several advantages over existing approaches.
-These include an explicitly dynamic approach, which can detect evolutionary contingencies and reciprocal causation, coupled with the enormous flexibility of the Bayesian implementation in the Stan programming language, allowing researchers to naturally account for latent factors, repeated measures, measurement error, missing data, and other real-life complexities of comparative data.
-The coevolve R package supports some of these complexities (such as missing data), as well as a reasonable diversity of distributions, facilitating the adoption of this method for most R users, while some researchers might prefer the flexibility of coding their models directly in Stan.
+These include an explicitly dynamic approach, which can detect evolutionary contingencies and reciprocal causation, coupled with the enormous flexibility of a Bayesian implementation in the Stan programming language, allowing researchers to naturally account for latent factors, repeated measures, measurement error, missing data, and other real-life complexities of comparative data.
+The coevolve R package supports some of these complexities (such as missing data), as well as a reasonable diversity of distributions, facilitating the adoption of this method for most R users.
+Researchers who might prefer the flexibility of coding their models directly in Stan, can build on the annotated code provided with our worked empirical examples.
 
 Despite several advances, our method is still subject to several limitations, some specific to GDPMs, and some more broadly applicable.
-In cases where traits experience correlated drift, GDPMs will struggle to separate this effect from cross-selection (EXPAND ON THIS, PERHAPS WITH A REAL BIOLOGICAL EXAMPLE?).
-GDPMs assume that cross-selection effects are linear; this assumption can be somewhat relaxed by using clade-specific parameters, though the effects will still be 'locally' linear (i.e.
+First, in cases where traits experience correlated drift, GDPMs will struggle to separate this effect from cross-selection (EXPAND ON THIS, PERHAPS WITH A REAL BIOLOGICAL EXAMPLE?).
+Second, GDPMs assume that cross-selection effects are linear; this assumption can be somewhat relaxed by using clade-specific parameters, though the effects will still be 'locally' linear (i.e.
 within a given clade).
-As with any Bayesian estimation, GDPMs can be computationally costly, though we note that the Hamiltonian Monte Carlo sampler of Stan is substantially more efficient than conventional MCMC samplers, typically requiring only a few thousand instead of millions of iterations; the models presented in this paper generally took minutes to hours to run.
-As with all comparative analyses, GDPMs are subject to concerns about data quality and bias; e.g.
+Third, as with any Bayesian estimation, GDPMs can be computationally costly, though we note that the Hamiltonian Monte Carlo sampler of Stan is substantially more efficient than conventional MCMC samplers, typically requiring only a few thousand instead of millions of iterations; all the models presented here, including the empirical applications, generally took minutes to hours to run.
+Fourth, as with all comparative analyses, GDPMs are subject to concerns about data quality and bias; e.g.
 that extant trait variation reflects past variation, that available phylogenies accurately reflect true evolutionary history, etc.
-We also note that since GDPMs work as a time series, inference or knowledge about ancestral states is particularly crucial, and relevant information should be incorporated into the models - for instance, Ringen et al [@Ringen_preprint] used priors for low complexity on the last common ancestor of all human societies; in this way, virtual archeology (or paleontology) can be enhanced by real archeology.
+We also note that since GDPMs explicitly work as a time series, inference or knowledge about ancestral states is particularly crucial, and relevant information should be incorporated into the models whenever possible - for instance, Ringen et al [@Ringen_preprint] used priors for low complexity on the last common ancestor of all human societies; in this way, virtual archeology (or paleontology) can be enhanced by real archeology.
 Lastly, we caution that GDPMs detect contingencies, and not causation per se, though this caveat applies to all approaches to causal inference from observational data [@Pearl2009; @Pagel1994].
